---
title: "Anderson_etal2025_code"
author: "Maggie Anderson"
date: "2025-10-31"
output: html_document
editor_options: 
  chunk_output_type: console
---

**Author(s):** Maggie Anderson
 
**Overview:** this R script evaluates the TeRaCON phenology data using a generalized linear mixed effects modeling approach. It also runs code and plots raw data and coefficients for supplementary analyses

[**Index**](#index)
--
  * [Setup](#setup): load in data/libraries and set working directory
  * [Data prep](#prep): prep data for phenology calculations 
  * [Phenology calculations](#phenoCalcs): phenology calculations; extract the first/last days of phenophase and length of phenophase time for each stage/subplot 
  * [Data wrangling](#wrangle): add treatment information, check for inconsistencies, fix variable names 
  * [Data structure](#str): investigate data structure
  * [Fixed effects correlation](#fixed_corr): look at correlation of fixed effects in data frame
  * [GDD & SPEI](#gdd_spei): extract SPEI and look at relationships between GDD and first dates of things 
  * [Climate data wrangling](#climate): climate data wrangling and plotting
  * [Correlation of response vars](#response_corr): how correlated are responses across phases, measures, and treatments? 
  * [Interannual variability](#yr_effects): interannual variability effects on phenology
  * [Model selection](#model_selection): selection for best-fit models, given interacting treatment effects
  * [Model checks](#check_models): graph residuals, address model complexity and overfitting
  * [Graph model predictions](#main_graphs): graph model means and SEs using predict()
  * [Graph all treatment interactions (models)](#all_trt_graphs): graph model coefficients, but include ALL non-sig. interactions 
  * [Climate info part 2](#gdd): calculate heat treatments effect on growing degree days
  * [Species-level GDD graphs](#sps_gdd)
  * [Species-level treatment effects](#sps_trt): main treatment effects (only) on individual species (regardless of year)
  * [Species-level year effects](#sps_yr): year effects on individual species (regardless of treatment) 
  * [Soil moisture graphs](#soil_m): visualize monthly soil moisture data during study
--

#### Setup: load in data/libraries and set working directory {#setup}
```{r, message=FALSE, warning=FALSE}
# clear workspace
rm(list = ls()) # don't run if you have objects in your env. that you want to save
graphics.off() 
  
# load in libraries
 library(AICcmodavg) # model selection via AICc
 library(car)        # test for multicollinearity
 library(caret)      # cross-validation of results
 library(corrplot)   # correlation plots
 library(data.table) # as.data.table()
 library(e1071)      # for checking skewness
 library(emmeans)    # post-hoc analyses
 library(ggpmisc)    # stat_poly_line()
 library(ggpubr)     # ggarrange for multiple plots
 library(ggthemes)   # nice-looking plots
 library(gridExtra)  # arrange plots
 library(lme4)       # models
 library(lmerTest)   # wrapper for models
 library(lubridate)  # coercing dates into workable formats
 library(MuMIn)      # streamlined model selection
 library(ncdf4)      # year conditions
 library(nlme)       # get DF approximations from models
 library(patchwork)  # arrange plots
 library(performance)# detecting overdispersion
 library(raster)     # year conditions
 library(rstatix)    # normality tests
 library(scales)     # extended graph capabilities
 library(shadowtext) # display significance codes on plots
 library(tidyverse)  # data wrangling, dplyr, ggplot2
 library(usdm)       # check for multicollinearity
 library(viridis)    # color scaling for ggplot

# set working directory (wherever you uploaded your data after downloading from GitHub)
# *your working directory here*

# Daily climate summary data from on-site weather station
weather <- read.csv("e080_DailyClimateSummary.csv")

# Soil moisture data from TeRaCON
sm <- read.csv("e141_SoilMoisture.csv")

# Load e141 TeRaCON phenology data (2012-2021)
df <- read.csv("TeRaCON_phenology_for_RM_210212.csv")

# Load e141 phenology data (2021-2022)
dfn <- read.csv("W2CONphenology2021-22.csv")

# Load e141 TeRaCON phenology data (2023)
df23 <- read.csv("W2CON_phenology_2023_COMPLETE.csv")

# Treatment-level data
dft <- read.csv("TeRaCON_treatments.csv")

################################################################################
# end setup
```
[Back to Index](#index)

#### Prep data for phenology calculations {#prep}
```{r}
# Change NAs to 0 
  df[is.na(df)] <- 0
  df[,8:25] <- lapply(df[,8:25], as.factor)

# Reformat to data table (otherwise code below doesn't run)
df <- as.data.table(df)

# fix misspelling of agrre
df$Species.observed <- as.factor(ifelse(df$Species.observed == "aggre", "agrre",df$Species.observed))
      df.gap <- df # save for later gap analysis

# Clean up 2021-2022 data
  # which species are in the 2021/22 dataset that aren't in 2012-19 dataset
  unique(df$Species.observed)
  unique(dfn$Species) # we added all the forb species in 2021 & 2022
  
# Replace spaces in 2023 df with periods in all column names
names(df23) <- gsub(" ", ".", names(df23))
# remove some unnecessary columns in df23
df23_corrected1 <- df23 %>% dplyr::select(-c(`Year.Day`, Sampling.., X,))
# rename species column
df23_corrected2 <- df23_corrected1 %>% rename(Species = Species.observed)
# Reorder columns so that Plot comes before Ring
df23_corrected <- df23_corrected2[, c(1, 2, 4, 3, 5:ncol(df23_corrected2))]

# Combine data from both datasets (collection protocols are the same)
  # 2021
  df21_22_23 <- bind_rows(dfn, df23_corrected) # formerly df23_partial
    
  dfn1 <- df21_22_23 %>% # replaced dfn with df21_22_23
    dplyr::rename(Species.observed = Species) %>%   # First, change "Species" column to "Species.observed"
    mutate(Species.observed = ifelse(Species.observed == "aggre", "agrre",Species.observed)) %>% # fix misspelling of agrre
    filter(Species.observed %in% c("agrre","luppe","solri","broin","lesca","schsc","andge","poapr")) %>% # trim species from 2021/22 dataset that aren't in 2012-19 dataset
    dplyr::select(-c(OpenFlowerCount_WholePlot))  %>% # remove this column (we're not using this for now)
    replace(is.na(.), 0) %>% # replace NAs with zero
    mutate(Ring = as.integer(Ring),
           Plot = as.integer(Plot),
           Species.observed = as.factor(Species.observed))
    
  # Add column to dfn for day of the year
  dfn1$Year.Day <- as.Date(dfn1$Date, "%m/%d/%Y")
  dfn1$Year.Day <- as.integer(format(as.POSIXct(dfn1$Year.Day,format='%m/%d/%Y'),format='%j'))
  
  # Keep last N characters in string to get date alone
  dfn1$Year <- as.integer(str_sub(dfn1$Date, start=-4))
  
  # Change all phenophase columns to factors (this is essential for the next chunk to run)
  dfn1[,5:22] <- lapply(dfn1[,5:22], as.factor)
  
# stack columns together
  df <- bind_rows(df, dfn1)
  
# Finish changes to original dataframe
  # Create unique column incorporating Year, Ring, Plot, Species
  df <- df %>%
    unite("Unique", c(Year, Ring, Plot, Species.observed))
  df$Unique <- as.factor(df$Unique)
  
# end prep
```
[Back to Index](#index)

#### Phenology calculations: extract the first/last days of phenophase and length of phenophase time for each stage/subplot {#phenoCalcs}
    # Contributors: Ben Tomhave (MN DOT) 
```{r}
# Group Df into many unique df one each for each unique value group
groupedDfList <- split(df, df$Unique)

possiblePhases <- list('Initial.Growth.subplot.1',
                       'Initial.Growth.subplot.2',
                       'Initial.Growth.subplot.3',
                       'Leaves.subplot.1',
                       'Leaves.subplot.2',
                       'Leaves.subplot.3',
                       'Flower.subplot.1',
                       'Flower.subplot.2',
                       'Flower.subplot.3',
                       'Open.Flowers.subplot.1',
                       'Open.Flowers.subplot.2',
                       'Open.Flowers.subplot.3',
                       'Fruits.subplot.1',
                       'Fruits.subplot.2',
                       'Fruits.subplot.3',
                       'Ripe.Fruits.subplot.1',
                       'Ripe.Fruits.subplot.2',
                       'Ripe.Fruits.subplot.3')

# Create blank output vectors
output_UniqueCharacter <- c()
output_StartDate <- c()
output_EndDate <- c()
output_Phase <- c()

# For each "Unique" df loop to find start and end dates
for (singleGroup in groupedDfList){
  singleGroup <- data.table(singleGroup)
  
  # For each phase, extract if two 1s in a row 
  for(Phase in possiblePhases) {
    
    # Identify start and end dates for specified column where column binary is 1 N times in a row
    outcome <- singleGroup[, if(.N > 1 & all(get(Phase) == 1)) .(starts = first(.I),
                                                                     ends = last(.I)),
                         by = rleid(get(Phase))] # Phase = string saved as variable, get converts the "Phase" variable back into a string
  
    # If there is data present in that column then output dates
    if (nrow(outcome) != 0){
      firstPatternDataOnly <- outcome[1, ] # Only Take first pattern occurrence if multiple
      # Get Start and End Dates
      startIndex <- as.numeric(firstPatternDataOnly[, starts])
      endIndex <- as.numeric(firstPatternDataOnly[, ends])
      startDate <- singleGroup[c(startIndex), Year.Day]
      endDate <- singleGroup[c(endIndex), Year.Day]

      
      # Get "Unique" Character Name
      uniqueVal <- as.character(singleGroup[c(endIndex), Unique])
      
      # Export Three Relevant Columns to Vectors to later be joined into df
      output_UniqueCharacter <- c(output_UniqueCharacter, uniqueVal)
      output_StartDate <- c(output_StartDate, startDate)
      output_EndDate <- c(output_EndDate, endDate)
      output_Phase <- c(output_Phase, Phase)
    }
    }
}

# # Combine three output lists above into a df
VALIDATION <- data.frame(output_UniqueCharacter, output_Phase, output_StartDate, output_EndDate)
# 
# # Separate Unique Into Multiple Columns and rename columns to match existing method
dff <- VALIDATION %>%
      dplyr::rename(Unique = output_UniqueCharacter)%>%
      dplyr::rename(start  = output_StartDate)%>%
      dplyr::rename(end    = output_EndDate)%>%
      dplyr::rename(Phase  = output_Phase) %>%
      separate(Unique, c("Year","Ring","Plot","Species.observed"), "_") %>%
      separate(Phase, c("Phase", "Subplot"), ".subplot.")

# Make new cols for phase length and phase midpoint
dff$length <- dff$end - dff$start
dff$midpt <- dff$length/2 + dff$start

# Validation: we can use the commented code below to test if our data has been compiled correctly
# TESTING
#  test <- data.table(df) %>%
#    filter(Unique == "2012_2_106_andge")
#  outcomeTest <- test[, if(.N > 1 & all(Leaves.subplot.2 == 1)) .(starts = first(.I),
#                                                                       ends = last(.I)),
#                           by = rleid(Leaves.subplot.2)][1, ]
#  
#  startIndex <- as.numeric(outcomeTest[, starts])
#  endIndex <- as.numeric(outcomeTest[, ends])
# # 
#  startDate <- test[c(startIndex), Year.Day]
#  endDate <- test[c(endIndex), Year.Day]
#  print(c(startDate, endDate))

# end phenoCalcs
```
[Back to Index](#index)

#### Data wrangling --add treatments, check for inconsistencies, fix variable names {#wrangle}
```{r}
# Take means across subplots and convert to long format
dff <- dff %>%
  group_by(Year, Ring, Plot, Species.observed, Phase) %>%
  dplyr::select(-Subplot) %>% # take this out so that we can summarise
      summarise_all(funs(mean(., na.rm = TRUE))) %>%
      pivot_longer(cols = start:midpt,
                   names_to = "Measure",
                   values_to = "YearDay") 

# Random fixes

# new column w/ grass vs. forb info (we don't use this in the analysis, but it's helpful for keeping track of species)
dff$Life.form <- ifelse(dff$Species.observed == "agrre" |
                          dff$Species.observed == "andge" |
                          dff$Species.observed == "broin" |
                          dff$Species.observed == "poapr" |
                          dff$Species.observed == "schsc", "Grass",
                        ifelse(dff$Species.observed == "luppe" |
                                 dff$Species.observed == "lesca" |
                                 dff$Species.observed == "solri", "Forb", NA))

dff[,2:3] <- lapply(dff[,2:3], as.integer) # convert class of Ring and Plot variables

dff <- left_join(dff, dft[,c(1:4,6,8)], by = c("Ring","Plot")) 

# fill blanks in water treatment column (should be H2O ambient)
dff$Water.Treatment <- ifelse(dff$Water.Treatment == "","H2Oamb",dff$Water.Treatment)

# create new column with all possible treatment combinations
dff$ClmTrt <- as.factor(ifelse(dff$Water.Treatment == "H2Oamb" & dff$Temp.Treatment == "HTamb", "Control",
                     ifelse(dff$Water.Treatment == "H2Oneg" & dff$Temp.Treatment == "HTamb","-Water",
                            ifelse(dff$Water.Treatment == "H2Oamb" & dff$Temp.Treatment == "HTelv" | dff$Plot == 226, "+Heat",
                                   ifelse(dff$Water.Treatment == "H2Oneg" & dff$Temp.Treatment == "HTelv","+Heat-Water",NA)))))

# Convert YearDay to an integer
dff$YearDay <- as.integer(dff$YearDay)
  dff.info <- dff %>% # save for later
      mutate(Plot = as.character(Plot),
             Ring = as.character(Ring))
  
dff.allPhases <- dff %>% # keep for looking and flower and fruit maturation
  filter(Year != 2012)

# apply filters to main dataframe
dff <- dff %>%
  filter(Phase != "Initial.Growth") %>%
  filter(Phase != "Flower") %>%
  filter(Phase != "Fruits") %>%
  mutate(Phase = ifelse(Phase == "Open.Flowers","Open flowers",
                         ifelse(Phase == "Ripe.Fruits","Ripe fruits",Phase))) 

# reorder factors for plotting
   dff$Measure <- factor(dff$Measure, levels = c("start","midpt","end","length"))
   dff$Phase <- factor(dff$Phase, levels = c("Leaves","Open flowers","Ripe fruits"))
   dff$ClmTrt <- factor(dff$ClmTrt, levels = c("Control","-Water","+Heat","+Heat-Water"))

length(unique(dff$Plot)) # 48 unique plots

# Take out midpt and end, since these are highly correlated with start and length
dff <- dff %>%
# Add experiment year to dataframe
  mutate(ExpYr = -(2011-as.numeric(Year))) %>%
  rename(N.Treatment = Ntreatment,
         CO2.Treatment = CO2Treatment) # rename some cols for consistency

# plot model coefficients (we don't use this in the analysis, but it's helpful for reference)
{      dff.mod <- dff %>%
        dplyr::mutate(ExpYrCat = as.factor(Year)) 
      # leaf start by species
          model.sps.s.le <- lmer(YearDay ~ ExpYrCat * Species.observed + (1|Ring), data = subset(dff.mod, CO2.Treatment == "AC" & N.Treatment == "AN" & Water.Treatment == "H2Oamb" & Temp.Treatment == "HTamb" & Measure == "start" & Phase == "Leaves"))
      # flower start by species
          model.sps.s.fl <- lmer(YearDay ~ ExpYrCat * Species.observed + (1|Ring), data = subset(dff.mod, CO2.Treatment == "AC" & N.Treatment == "AN" & Water.Treatment == "H2Oamb" & Temp.Treatment == "HTamb" & Measure == "start" & Phase == "Open flowers"))
      # fruit start by species
          model.sps.s.fr <- lmer(YearDay ~ ExpYrCat * Species.observed + (1|Ring), data = subset(dff.mod, CO2.Treatment == "AC" & N.Treatment == "AN" & Water.Treatment == "H2Oamb" & Temp.Treatment == "HTamb" & Measure == "start" & Phase == "Ripe fruits"))
          
      # leaf end by species
          model.sps.l.le <- lmer(YearDay ~ ExpYrCat * Species.observed + (1|Ring), data = subset(dff.mod, CO2.Treatment == "AC" & N.Treatment == "AN" & Water.Treatment == "H2Oamb" & Temp.Treatment == "HTamb" & Measure == "end" & Phase == "Leaves"))
      # flower end by species
          model.sps.l.fl <- lmer(YearDay ~ ExpYrCat * Species.observed + (1|Ring), data = subset(dff.mod, CO2.Treatment == "AC" & N.Treatment == "AN" & Water.Treatment == "H2Oamb" & Temp.Treatment == "HTamb" & Measure == "end" & Phase == "Open flowers"))
      # fruit end by species
          model.sps.l.fr <- lmer(YearDay ~ ExpYrCat * Species.observed + (1|Ring), data = subset(dff.mod, CO2.Treatment == "AC" & N.Treatment == "AN" & Water.Treatment == "H2Oamb" & Temp.Treatment == "HTamb" & Measure == "end" & Phase == "Ripe fruits"))}

# Function to extract emmeans and SE from models
extract_emmeans <- function(model) {
  emmeans_output <- emmeans(model, ~ ExpYrCat * Species.observed)
  summary_output <- summary(emmeans_output)
  return(summary_output)
}

# Models
models <- list(
  model.sps.s.le = model.sps.s.le,
  model.sps.s.fl = model.sps.s.fl,
  model.sps.s.fr = model.sps.s.fr,
  model.sps.l.le = model.sps.l.le,
  model.sps.l.fl = model.sps.l.fl,
  model.sps.l.fr = model.sps.l.fr
)

# Extract emmeans and SE for each model
emmeans_list <- lapply(models, extract_emmeans)

# Combine results into a data frame
results <- do.call(rbind, lapply(names(emmeans_list), function(name) {
  data.frame(Model = name, emmeans_list[[name]])
}))

# Custom labels for the models
model_labels <- c(
  model.sps.s.le = "Leaf Start",
  model.sps.s.fl = "Flower Start",
  model.sps.s.fr = "Fruit Start",
  model.sps.l.le = "Leaf End",
  model.sps.l.fl = "Flower End",
  model.sps.l.fr = "Fruit End"
)

# Desired order of the model levels
model_order <- c("model.sps.s.le", "model.sps.s.fl", "model.sps.s.fr", "model.sps.l.le", "model.sps.l.fl", "model.sps.l.fr")

# Reorder the levels of the Model variable
results$Model <- factor(results$Model, levels = model_order)

# Custom vector for the new facet labels
species_labels <- c(
  agrre = "italic('E. repens')",
  andge = "italic('A. gerardii')",
  broin = "italic('B. inermis')",
  lesca = "italic('L. capitata')",
  luppe = "italic('L. perennis')",
  poapr = "italic('P. pratensis')",
  schsc = "italic('S. scoparium')"
)
# Sort the labels alphabetically
species_labels <- species_labels[order(names(species_labels))]


# Plotting with custom facet labels
# Update the ggplot code
ggplot(results, aes(y = ExpYrCat, x = emmean, xmin = emmean - SE, xmax = emmean + SE, color = ExpYrCat)) +
  geom_pointrange(size = 0.15) +
  facet_grid(Species.observed ~ Model, labeller = labeller(Species.observed = as_labeller(species_labels, label_parsed), Model = as_labeller(model_labels)), scales = "free") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Model", x = "Day of the Year (start) / (end)", y = element_blank(), color = "Year") +
  scale_color_viridis(option = "H", discrete = TRUE) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12),
    axis.title = element_text(size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text = element_text(size = 8),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  )

df <- as.data.frame(dff) # new dataframe version

# means and sds across all species
sum.tbl <- df %>%
  group_by(ExpYr,Phase,Measure,Life.form,Water.Treatment,Temp.Treatment,CO2.Treatment,N.Treatment) %>% # means and sds across all species
  summarise(n=n(),
            mean=mean(YearDay),
            sd=sd(YearDay)) # MIA: added ExpYr

df <- df %>%
  mutate(Phase = ifelse(Phase == "Open flowers", "OpenFlowers",
                        ifelse(Phase == "Ripe fruits", "RipeFruits", 
                               ifelse(Phase == "Leaves", "Leaves", NA)))) # change reproductive phase names for coding ease

# here I make a column called "trait" which is a combination of phase and life form (this is useful later)
df2 <- df %>% # used to be df2
  pivot_wider(names_from = Measure,values_from = YearDay) %>% mutate(trait = paste(Life.form,Phase, sep = "_")) %>% 

# make sure only the 48 treated plots are used
  drop_na(Temp.Treatment, Water.Treatment, CO2.Treatment, N.Treatment)

unique_plots <- data.frame(sort(unique(df2$Plot))); dim(unique_plots)

# random sidequest: fix labels
df2$CO2.Treatment <- factor(df2$CO2.Treatment, levels = c("AC", "EC"), 
                  labels = c("ambient CO2", "+CO2"))
df2$N.Treatment <- factor(df2$N.Treatment, levels = c("AN", "EN"),
                  labels = c("ambient N", "+N")
                  )

# Add categorical variable for experimental ring, year, & plot
df2 <- df2 %>%
  dplyr::mutate(ExpYrCat = as.factor(ExpYr),
                Ring = as.factor(Ring),
                Plot = as.factor(Plot))

 # end wrangle
```
[Back to Index](#index)

#### Investigate data structure (response variation/data normality) {#str}
```{r, message=FALSE, warning=FALSE, fig.width=8, fig.height=8}
# remember that I elected to throw out RipeFruits based on the correlation matrix generated in Part 1
# start   
    ggplot(df2, aes(x = start)) +
      geom_density(aes(color = Phase, fill = Phase), 
                     position = "identity", alpha = 0.4) +
      scale_color_brewer(palette = "Set1", direction = -1) +
      scale_fill_brewer(palette = "Set1", direction = -1) +
      labs(x = "Start of phase (DOY)", subtitle = "Phase start distribution by species") +
      facet_wrap(~Species.observed, scales = "free") +
      theme_few() 
# end  
    ggplot(df2, aes(x = end)) +
      geom_density(aes(color = Phase, fill = Phase), 
                     position = "identity", alpha = 0.4) +
      scale_color_brewer(palette = "Set1", direction = -1) +
      scale_fill_brewer(palette = "Set1", direction = -1) +
      labs(x = "End of phase (days)", subtitle = "Phase end distribution by species") +
      facet_wrap(~Species.observed) +
      theme_few()
    
# test for normality within groups
    norm1 <- df2 %>%
      group_by(Phase, Species.observed) %>%
      shapiro_test(start,end,midpt) %>%
      mutate(action = ifelse(p < 0.05, "non-normal",""));print(norm1,n=32) # p-value > 0.05 would imply that the distribution of the data are not significantly different from normal distribution
      # ^^ most of these data categories could benefit from data transformations
    
    norm2 <- df2 %>%
      group_by(Phase, Species.observed) %>%
      mutate(start.log = log(start),
             end.log = log(end),
             midpt.log = log(midpt)) %>%
      shapiro_test(start.log,end.log) %>%
      mutate(action = ifelse(p < 0.05, "non-normal",""));print(norm2,n=32)
    # log transformations do not change the normality of the data distributions overly much...
    
# end structure investigation
```
[Back to Index](#index)

#### look at correlation of fixed effects in data frame {#fixed_corr}
It is important to understand which variables are most correlated with each other. This will help us decide whether to keep them in the model. In general, I am following the guideline that a correlation is considered strong in ecology when |r| > 0.7. 
 * Source: David N. Chapter 6—Selection of Variables and Factor Derivation. Commercial data mining: Processing, analysis and modeling for predictive analytics projects. 2014:79-104.
```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}
# make dataframe for correlation matrix
# correlation regardless of species (otherwise we have NAs in dataframe)
# ONLY CONTROL PLOTS
df.corr <- df2 %>% # remove treatment/ring information and a redundant column for now 
  as.data.frame() %>%
  dplyr::filter(Water.Treatment == "H2Oamb" &
                Temp.Treatment == "HTamb" &
                CO2.Treatment == "ambient CO2" &
                N.Treatment == "ambient N") %>% # only use data from control plots
  dplyr::select(Ring, Plot, Species.observed, Phase, ExpYr, start, end) %>%
  mutate(start = as.numeric(start),
         end = as.numeric(end)) %>%
  pivot_longer(cols = start:end, names_to = "Measure",values_to = "value") %>%
  dplyr::group_by(ExpYr, Ring, Plot, Species.observed) %>%
  pivot_wider(names_from = c(Phase,Measure),values_from = value, values_fn = mean) %>%
  ungroup() %>%
  dplyr::select(-c(ExpYr, Ring, Plot, Species.observed)) # now we need to remove our grouping columns (except ExpYr because this is a ifixed effect) before we test for correlations
  
# calculate correlation matrix
cor_matrix <- cor(df.corr, use='pairwise.complete.obs') 

# Calculate the correlation matrix
    # Here, "pairwise.complete.obs" computes the correlation or covariance between each pair of variables using only the complete pairs of observations. This means that if there are missing values for some observations, those pairs will be excluded from the calculation. However, this can lead to incomplete correlation matrices, and in some cases, it can result in NA entries if there are no complete pairs for a pair of variables.

corrplot(cor_matrix, type = "upper",
         method = "square",
         addCoef.col = "black",
         tl.col = "black", tl.srt = 45,
         main = "
         Correlation of Phases and Measures (control plots, all years)")

# To be sure, let's test multicollinearity using the VIF function
# Generally if VIF > 4, then we assume that the variables are correlated
    df.mcl <- df2 %>%
      #filter(Life.form == "Forb") %>%
      dplyr::select(-c(trait,midpt,Water.Treatment,Temp.Treatment,CO2.Treatment,N.Treatment,Life.form)) %>% # we don't need this column for vif
      group_by(ExpYr,Ring,Plot,Species.observed) %>% # dataframe into wide format
      pivot_wider(names_from = Phase, values_from = c(end,start)) %>%
      as.data.frame()

    df2.backup <- df2 # maybe useful later
    
# end fixed effects correlation
```
[Back to Index](#index)

#### extract SPEI and look at relationships between GDD and first dates of things {#gdd_spei}
```{r}
##############
# SPEI data
##############
    
    # load in the new SPEI dataframe
    # Use nc_open to read the data into a data structure I called nc_data. Print the metadata about this file to a text file.
    our_nc_data <- nc_open("spei04_45.5_-93.2.nc") #took out: G:/My Drive/Chapter2_phenology/Rcode&data/
    print(our_nc_data)
    
    attributes(our_nc_data$var)
    attributes(our_nc_data$dim)
    
    lat <- ncvar_get(our_nc_data, "lat") # need data for 45.25
    nlat <- dim(lat) #to check it matches the metadata: 23
    lon <- ncvar_get(our_nc_data, "lon") # need data for -93.25
    nlon <- dim(lon) #to check (is this number of grid cells?)
    nlon
    
    # Find the indices for the desired latitude and longitude
    lat_index <- which(lat == 45.25) 
    lon_index <- which(lon == -93.25) 
    
    # Extract the "spei" variable
    spei_data <- ncvar_get(our_nc_data, "spei")
    fill_value <- ncatt_get(our_nc_data, "spei", "_FillValue")$value
    
    # Extract the "spei" data for the specific location
    spei_specific <- spei_data[lon_index, lat_index, ]
    spei_specific[spei_specific == fill_value] <- NA
    
    # Extract the time variable and its units
    time <- ncvar_get(our_nc_data, "time")
    tunits <- ncatt_get(our_nc_data, "time", "units")$value
    print(tunits)
    
    # Convert the time variable to a readable date format
    # The time units are "days since 1900-1-1"
    time_origin <- as.Date("1900-01-01")
    time_dates <- as.Date(time, origin = time_origin)
    
    # Combine the time and "spei" values into a dataframe
    spei_df <- data.frame(
      Date = time_dates,
      SPEI = spei_specific
    )
    
    # Print the dataframe
    print(spei_df)
    
    
    # Add separate columns for year, month, and day
    spei_df <- spei_df %>%
      mutate(
        Year = as.numeric(format(Date, "%Y")),
        Month = as.numeric(format(Date, "%m")),
        Day = as.numeric(format(Date, "%d"))
      ) %>%
      filter(SPEI != is.na(SPEI)) %>%
      # filter to only get years we're interested in
      filter(Year %in% c(2012, 2013, 2014, 2015, 2016, 2017, 2018, 2019, 2020, 2021, 2022, 2023))
    
    # Close the NetCDF file
    nc_close(our_nc_data)
    
    # wrangle
    spei1.1 <- spei_df %>% 
      data.frame() %>%
      dplyr::filter(Month == 8) # Month formerly set to "Aug"
    
    spei1.1$SPEI_4cat <- "Normal"
    qnorm(c(0.1,0.25,0.75,0.9)) #norm function provides the quantile of the normal distribution at a specified cumulative density (Isbell et al. 2015, Nature)
    
    spei1.1$SPEI_4cat[spei1.1$SPEI >= 1.2815516] <- "Extremely wet" 
    spei1.1$SPEI_4cat[spei1.1$SPEI >= 0.6744898 & spei1.1$SPEI < 1.2815516] <- "Wet"
    spei1.1$SPEI_4cat[spei1.1$SPEI <= -1.2815516] <- "Extremely dry"
    spei1.1$SPEI_4cat[spei1.1$SPEI < -0.6744898 & spei1.1$SPEI > -1.2815516] <- "Dry"
    
    unique(spei1.1$SPEI_4cat) # simplify these categories
    spei2 <- spei1.1 %>%
      mutate(status = ifelse(SPEI_4cat == "Normal","Normal",
                             ifelse(SPEI_4cat == "Dry" | SPEI_4cat == "Extremely dry","Dry",
                                    ifelse(SPEI_4cat == "Wet" | SPEI_4cat == "Extremely wet","Wet",NA))),
             severity = ifelse(SPEI_4cat == "Normal","Normal",
                               ifelse(SPEI_4cat == "Moderately wet" | SPEI_4cat == "Moderately dry","Moderate",
                                      ifelse(SPEI_4cat == "Extremely wet" | SPEI_4cat == "Extremely dry","Extreme","None")))) %>%
      rename(SPEI04 = SPEI) %>%
      mutate(Year = as.character(Year)) %>%
      rename(event = status)
    
    # look at data
    head(spei2) 
# end GDD prep and SPEI extraction
```
[Back to Index](#index)

#### climate data wrangling and plotting {#climate}
```{r}
head(weather)

# Convert the DATE column to Date format
weather <- weather %>%
  mutate(Date = as.Date(Date, format = "%m/%d/%Y"))

# Exclude the year 2019 from the dataset
weather_filtered <- weather %>%
  filter(year(Date) != 2020) %>%
  filter(year(Date) > 2011 & year(Date) < 2024) %>%
  # change temp to degrees C and precip to mm
  dplyr::mutate(MaxTemp.degC. = (MaxTemp.degF.- 32)*5/9,
                MinTemp.degC. = (MinTemp.degF.- 32)*5/9,
                Precip.mm. = Precip.inches.*25.4)

# Define a function to calculate GDD
calculate_gdd <- function(MinTemp.degC., MaxTemp.degC., base_temp = 10) {
  MinTemp.degC. <- ifelse(MinTemp.degC. < base_temp, base_temp, MinTemp.degC.)
  MaxTemp.degC. <- ifelse(MaxTemp.degC. < base_temp, base_temp, MaxTemp.degC.)
  gdd <- ((MaxTemp.degC. + MinTemp.degC.) / 2) - base_temp
  return(max(gdd, 0))
}

# Calculate GDD and GDD with temperature treatment for each day
weather_gdd <- weather_filtered %>%
  mutate(GDD = mapply(calculate_gdd, MinTemp.degC., MaxTemp.degC.),
         ) # precipitation treatment

# 1. Create a dataset with cumulative GDD and total precipitation by year
cumulative_gdd_year <- weather_gdd %>%
  group_by(Year = year(Date)) %>%
  summarize(Cumulative_GDD = sum(GDD, na.rm = TRUE),
            Total_Precipitation_mm = sum(Precip.mm., na.rm = TRUE) 
            ) %>%
  arrange(Year) %>%
  dplyr::mutate(ExpYrCat = as.factor(-(2011-Year)))


# add to df2
df2 <- df2 %>%
  left_join(., spei2[,c("Year", "SPEI04","event")], by = "Year")

# end cliamte data wrangle
```
[Back to Index](#index)

#### How correlated are responses across phases, measures, and treatments? {#response_corr}
I want to spend time digging thoroughly into the correlations between the start and end of leafing, flowering, and fruiting (think a correlation figure with 6 panels). Rather than seeing these correlated variables as an inconvenience, I want to lean into them.*Note: this section is not included in the main analyses, but provides useful context*
```{r}
# first, make data subsets based on these troublesome models:
flower_start <- df2 %>% filter(Phase == "OpenFlowers") %>% drop_na(start) %>% mutate(ExpYrCat = as.factor(ExpYr))
    flower_start_mod <- lmer(start ~ Temp.Treatment * Water.Treatment * CO2.Treatment * N.Treatment - (Temp.Treatment: Water.Treatment:CO2.Treatment:N.Treatment) - (Temp.Treatment: Water.Treatment:CO2.Treatment) - (Temp.Treatment: Water.Treatment:N.Treatment) - (Temp.Treatment: CO2.Treatment:N.Treatment) - (Water.Treatment:CO2.Treatment:N.Treatment)+ (1|ExpYrCat) + (1|Ring) +  (1|Species.observed), data = flower_start)
    
flower_end <- df2 %>% filter(Phase == "OpenFlowers") %>% drop_na(end) %>% mutate(ExpYrCat = as.factor(ExpYr))
        flower_end_mod <- lmer(end ~ Temp.Treatment * Water.Treatment * CO2.Treatment * N.Treatment- (Temp.Treatment: Water.Treatment:CO2.Treatment:N.Treatment) - (Temp.Treatment: Water.Treatment:CO2.Treatment) - (Temp.Treatment: Water.Treatment:N.Treatment) - (Temp.Treatment: CO2.Treatment:N.Treatment) - (Water.Treatment:CO2.Treatment:N.Treatment) + (1|ExpYrCat) + (1|Ring/Plot) +  (1|Species.observed), data = flower_end)
        
fruit_end <- df2 %>% filter(Phase == "RipeFruits") %>% drop_na(end) %>% mutate(ExpYrCat = as.factor(ExpYr))
        fruit_end_mod <- lmer(end ~ Temp.Treatment * Water.Treatment * CO2.Treatment * N.Treatment + (1+SPEI04|ExpYrCat) + (1|Ring/Plot) + (1|Plot/Species.observed), data = fruit_end)

# 1) Check for unbalanced species/treatment combinations
    print("flower_start")
    table(flower_start$Temp.Treatment, flower_start$Water.Treatment, flower_start$CO2.Treatment, flower_start$N.Treatment, flower_start$Species.observed) # no major trt holes (lowest is 52), but several sps holes (7)
    
    print("flower_end")
    table(flower_end$Temp.Treatment, flower_end$Water.Treatment, flower_end$CO2.Treatment, flower_end$N.Treatment, flower_end$Species.observed) # ALSO no major holes (lowest is 52), but lots of sps holes (17)
    
    print("fruit_end")
    table(fruit_end$Temp.Treatment, fruit_end$Water.Treatment, fruit_end$CO2.Treatment, fruit_end$N.Treatment, fruit_end$Species.observed) # ALSO no major holes (lowest is 63), but lots of sps holes (15)
    
    
# 2) Check for SPEI04 imbalances
      # Eight levels are minimal but often sufficient for estimating a random slope. However, this depends on how balanced these levels are across ExpYrCat. 
      # If one level of SPEI04 dominates or corresponds to only one level of ExpYrCat, it can destabilize the model.
    table(flower_start$SPEI04, flower_start$ExpYrCat)
    table(flower_end$SPEI04, flower_end$ExpYrCat)
    table(fruit_end$SPEI04, fruit_end$ExpYrCat) # yeah this is ALL super unevenly distributed

# 3) Check for convergence or singularity in models
    print("flower_start")
    summary(flower_start_mod)  # Look for "boundary (singular) fit" warnings.
    isSingular(flower_start_mod, tol = 1e-05)  # TRUE if singular.  
    
    print("flower_end")
    summary(flower_end_mod)  # Look for "boundary (singular) fit" warnings.
    isSingular(flower_end_mod, tol = 1e-05)  # TRUE if singular.
    
    print("fruit_end")
    summary(fruit_end_mod)  # Look for "boundary (singular) fit" warnings.
    isSingular(fruit_end_mod, tol = 1e-05)  # TRUE if singular.

    # both end models have a singular fit
    
    # If singularity occurs, I need to consider simplifying random effects. For example, remove the random slope (1 + SPEI04 | ExpYrCat) or collapse levels of random effects like Species.observed.
    
# 4) To detect issues with Species.observed decline, create a cross-tabulation of species by year:
    print("flower_start")
    table(flower_start$Species.observed, flower_start$ExpYrCat)
    print("flower_end")
    table(flower_end$Species.observed, flower_end$ExpYrCat)
    print("fruit_end")
    table(fruit_end$Species.observed, fruit_end$ExpYrCat)
    
# 5) Multicollinearity might inflate DoF. Use variance inflation factors (VIFs) to check:
    library(car)
    vif(lmer(start ~ Temp.Treatment * Water.Treatment * CO2.Treatment * N.Treatment - (Temp.Treatment: Water.Treatment:CO2.Treatment:N.Treatment) - (Temp.Treatment: Water.Treatment:CO2.Treatment) - (Temp.Treatment: Water.Treatment:N.Treatment) - (Temp.Treatment: CO2.Treatment:N.Treatment) - (Water.Treatment:CO2.Treatment:N.Treatment) + (1 + SPEI04 | ExpYrCat) + (1|Ring) +  (1|Species.observed), data = flower_start))
    vif(lmer(end ~ Temp.Treatment * Water.Treatment * CO2.Treatment * N.Treatment - (Temp.Treatment: Water.Treatment:CO2.Treatment:N.Treatment) - (Temp.Treatment: Water.Treatment:CO2.Treatment) - (Temp.Treatment: Water.Treatment:N.Treatment) - (Temp.Treatment: CO2.Treatment:N.Treatment) - (Water.Treatment:CO2.Treatment:N.Treatment) + (1 + SPEI04 | ExpYrCat) + (1|Ring) +  (1|Species.observed), data = flower_end))
    vif(lmer(end ~ Temp.Treatment * Water.Treatment * CO2.Treatment * N.Treatment - (Temp.Treatment: Water.Treatment:CO2.Treatment:N.Treatment) - (Temp.Treatment: Water.Treatment:CO2.Treatment) - (Temp.Treatment: Water.Treatment:N.Treatment) - (Temp.Treatment: CO2.Treatment:N.Treatment) - (Water.Treatment:CO2.Treatment:N.Treatment) + (1 + SPEI04 | ExpYrCat) + (1|Ring) +  (1|Species.observed), data = fruit_end))
    
# 6) Remember that missing levels in higher-order interactions can inflate DoF due to overparameterization. Check which levels are missing:
    print("flower_start")
    xtabs(~ Temp.Treatment + Water.Treatment + CO2.Treatment + N.Treatment + ExpYrCat + Species.observed, data = flower_start)
    print("flower_end")
    xtabs(~ Temp.Treatment + Water.Treatment + CO2.Treatment + N.Treatment + ExpYrCat + Species.observed, data = flower_end)
    print("fruit_end")
    xtabs(~ Temp.Treatment + Water.Treatment + CO2.Treatment + N.Treatment + ExpYrCat + Species.observed, data = fruit_end)

```

#### check the type, distribution, and skewness of response variables {#variable_checks}
# How do we decide what type of model to use?
# transform end varible 
```{r}
# Check if the data is count-based or continuous
is_count_data_start <- all(df2$start == floor(df2$start)) # TRUE if all values are integers
is_count_data_end <- all(df2$end == floor(df2$end))
cat("Start variable is count data:", is_count_data_start, "\n")
cat("End variable is count data:", is_count_data_end, "\n") 
    # it's count data, so we this limits the types of distributions we're able to use

# Histogram to inspect distribution
par(mfrow = c(1,2))
hist(df2$start, breaks = 15, col = "skyblue", main = "Start Variable", xlab = "start")
hist(df2$end, breaks = 15, col = "lightcoral", main = "End Variable", xlab = "end")
par(mfrow = c(1,1))

# Shapiro-Wilk test for normality (sensitive for larger datasets)
shapiro.test(df2$start[1:500])  # Subset to 500 to avoid errors for large datasets
shapiro.test(df2$end[1:500])
# If p < 0.05, the data significantly deviates from normality, favoring glmer().
# p-values < 2.2e-16, confirming non-normality.

# Quantify skewness
cat("Skewness (Start):", skewness(df2$start), "\n") # positive, mild but not extreme
cat("Skewness (End):", skewness(df2$end), "\n") # negative, more concerning
# Interpretation:
    # Right-skewed: Positive skewness = Poisson, Gamma, or Negative Binomial.
    # Left-skewed: Negative skewness = Consider zero-inflated or transformation-based models.

# both variables are non-normal, meaning that a glmer is a better option than an lmer, since glmer can better account for different data distributions
# HOWEVER, there are not a lot of good ways to handle the left-skewed data that we see with the "end" response. That said, we can transform the "end" variable relative to the last day of the year instead of the first to address this issue:
df2$end_rev <- 1*(365- df2$end)
hist(df2$end_rev, breaks = 20, col = "tomato1", main = "End Variable (relative to Dec. 31st)", xlab = "end")
df2$sqrt_end_rev <- sqrt(df2$end_rev + 1)  # Add 1 to avoid log(0)
df2$sqrt_end_rev <- sqrt(df2$end_rev + 1)  # Add 1 to avoid log(0)
hist(df2$sqrt_end_rev, breaks = 20, col = "cornflowerblue", main = "End Variable (relative to Dec. 31st)", xlab = "end")

# Response Variable Nature: Both start and end are count data, confirmed using:
all(df2$start == floor(df2$start))
all(df2$end == floor(df2$end))
# end response variable checks 
```
[Back to Index](#index)

#### interannual variability effects on phenology {#yr_effects}
*Note: this chunk takes a long time to run*
```{r}
# start of leaves
yr_effects_start_leaves <- glmer(start ~ Temp.Treatment*ExpYrCat + Water.Treatment*ExpYrCat + CO2.Treatment*ExpYrCat + N.Treatment*ExpYrCat + (1|Ring:Plot:Species.observed), data = subset(df2, Phase == "Leaves"), family = poisson)
            check_overdispersion(yr_effects_start_leaves) # good
            car::vif(yr_effects_start_leaves) # good
            anova(yr_effects_start_leaves) 
            Anova(yr_effects_start_leaves) # massive effect of year.
            emms <- emmeans(yr_effects_start_leaves, ~ Temp.Treatment | ExpYrCat);pairs(emms) 
            # lme for DF approximation
            yr_start_leaves_preds_lme <- lme(
                  fixed = start ~ Temp.Treatment*ExpYrCat + Water.Treatment*ExpYrCat + CO2.Treatment*ExpYrCat + N.Treatment*ExpYrCat,
                  random = list(Species.observed = ~1, Ring = ~1 | Plot),
                  data = subset(df2, Phase == "Leaves")
                );anova(yr_start_leaves_preds_lme)
            
# start of flowers     
yr_effects_start_flowers <- glmer(start ~ Temp.Treatment*ExpYrCat + Water.Treatment*ExpYrCat + CO2.Treatment*ExpYrCat + N.Treatment*ExpYrCat + (1|Ring:Plot:Species.observed), data = subset(df2, Phase == "OpenFlowers"), family = poisson)
            check_overdispersion(yr_effects_start_flowers) # good
            car::vif(yr_effects_start_flowers) # good
            anova(yr_effects_start_flowers) 
            Anova(yr_effects_start_flowers) # massive year effect
            # lme for DF approximation
            yr_start_flowers_preds_lme <- lme(
                  fixed = start ~ Temp.Treatment*ExpYrCat + Water.Treatment*ExpYrCat + CO2.Treatment*ExpYrCat + N.Treatment*ExpYrCat,
                  random = list(Species.observed = ~1, Ring = ~1 | Plot),
                  data = subset(df2, Phase == "OpenFlowers")
                );anova(yr_start_flowers_preds_lme)

# start of fruits          
yr_effects_start_fruits <- glmer(start ~ Temp.Treatment*ExpYrCat + Water.Treatment*ExpYrCat + CO2.Treatment*ExpYrCat + N.Treatment*ExpYrCat + (1|Ring:Plot:Species.observed), data = subset(df2, Phase == "RipeFruits"), family = poisson)
            check_overdispersion(yr_effects_start_fruits) # good
            car::vif(yr_effects_start_fruits) # good 
            anova(yr_effects_start_fruits) 
            Anova(yr_effects_start_fruits)
            # lme for DF approximation
            yr_start_fruits_preds_lme <- lme(
                  fixed = start ~ Temp.Treatment*ExpYrCat + Water.Treatment*ExpYrCat + CO2.Treatment*ExpYrCat + N.Treatment*ExpYrCat,
                  random = list(Species.observed = ~1, Ring = ~1 | Plot),
                  data = subset(df2, Phase == "RipeFruits")
                );anova(yr_start_fruits_preds_lme)
            
# end of leaves
yr_effects_end_leaves <- glmer(sqrt_end_rev ~ Temp.Treatment*ExpYrCat + Water.Treatment*ExpYrCat + CO2.Treatment*ExpYrCat + N.Treatment*ExpYrCat + (1|Ring:Plot:Species.observed), data = subset(df2, Phase == "Leaves"), family = Gamma(link = "log")) 
            car::vif(yr_effects_end_leaves) 
            anova(yr_effects_end_leaves) 
            Anova(yr_effects_end_leaves) 
            # lme for DF approximation
            yr_end_leaves_preds_lme <- lme(
                  fixed = sqrt_end_rev ~ Temp.Treatment*ExpYrCat + Water.Treatment*ExpYrCat + CO2.Treatment*ExpYrCat + N.Treatment*ExpYrCat,
                  random = list(Species.observed = ~1, Ring = ~1 | Plot),
                  data = subset(df2, Phase == "Leaves")
                );anova(yr_end_leaves_preds_lme)
# end of flowers
yr_effects_end_flowers <- glmer(sqrt_end_rev ~ Temp.Treatment*ExpYrCat + Water.Treatment*ExpYrCat + CO2.Treatment*ExpYrCat + N.Treatment*ExpYrCat + (1|Ring:Plot:Species.observed), data = subset(df2, Phase == "OpenFlowers"), family = Gamma(link = "log")) 
            car::vif(yr_effects_end_flowers) 
            anova(yr_effects_end_flowers) 
            Anova(yr_effects_end_flowers) 
            # lme for DF approximation
            yr_end_flowers_preds_lme <- lme(
                  fixed = sqrt_end_rev ~ Temp.Treatment*ExpYrCat + Water.Treatment*ExpYrCat + CO2.Treatment*ExpYrCat + N.Treatment*ExpYrCat,
                  random = list(Species.observed = ~1, Ring = ~1 | Plot),
                  data = subset(df2, Phase == "OpenFlowers")
                );anova(yr_end_flowers_preds_lme)
# end of fruits
yr_effects_end_fruits <- glmer(sqrt_end_rev ~ Temp.Treatment*ExpYrCat + Water.Treatment*ExpYrCat + CO2.Treatment*ExpYrCat + N.Treatment*ExpYrCat + (1|Ring:Plot:Species.observed), data = subset(df2, Phase == "RipeFruits"), family = Gamma(link = "log")) 
            car::vif(yr_effects_end_flowers) 
            anova(yr_effects_end_flowers) # effect of elevated CO2 despite year
            Anova(yr_effects_end_flowers) 
            # lme for DF approximation
            yr_end_fruits_preds_lme <- lme(
                  fixed = sqrt_end_rev ~ Temp.Treatment*ExpYrCat + Water.Treatment*ExpYrCat + CO2.Treatment*ExpYrCat + N.Treatment*ExpYrCat,
                  random = list(Species.observed = ~1, Ring = ~1 | Plot),
                  data = subset(df2, Phase == "RipeFruits")
                );anova(yr_end_fruits_preds_lme)
            
# visualize residuals
    par(mfrow = c(2,3))
    plot(yr_effects_start_leaves, main = "Residuals: Start/Leaves + Year, Poisson")
    plot(yr_effects_start_flowers, main = "Residuals: Start/Flowers + Year, Poisson")
    plot(yr_effects_start_fruits, main = "Residuals: Start/Fruits + Year, Poisson")
    
    plot(yr_effects_end_leaves, main = "Residuals: End/Leaves + Year, Gamma")
    plot(yr_effects_end_flowers, main = "Residuals: End/Flowers + Year, Gamma")
    plot(yr_effects_end_fruits, main = "Residuals: End/Fruits + Year, Gamma")
    par(mfrow = c(1,1))

# EXTRACT AND PLOT TREATMENT EFFECTS
get_treatment_effects <- function(model, measure, phase) {
  # make newdat (same structure for all models)
  newdat <- expand.grid(
    Temp.Treatment  = c("HTamb", "HTelv"),
    Water.Treatment = "H2Oamb",
    CO2.Treatment   = "ambient CO2",
    N.Treatment     = "ambient N",
    ExpYrCat        = levels(df2$ExpYrCat)
  ) %>% mutate(Treatment_Combo = "Temp") %>%
    
    bind_rows(expand.grid(
      Temp.Treatment  = "HTamb",
      Water.Treatment = c("H2Oamb", "H2Oneg"),
      CO2.Treatment   = "ambient CO2",
      N.Treatment     = "ambient N",
      ExpYrCat        = levels(df2$ExpYrCat)
    ) %>% mutate(Treatment_Combo = "Water")) %>%
    
    bind_rows(expand.grid(
      Temp.Treatment  = "HTamb",
      Water.Treatment = "H2Oamb",
      CO2.Treatment   = c("ambient CO2", "+CO2"),
      N.Treatment     = "ambient N",
      ExpYrCat        = levels(df2$ExpYrCat)
    ) %>% mutate(Treatment_Combo = "CO2")) %>%
    
    bind_rows(expand.grid(
      Temp.Treatment  = "HTamb",
      Water.Treatment = "H2Oamb",
      CO2.Treatment   = "ambient CO2",
      N.Treatment     = c("ambient N", "+N"),
      ExpYrCat        = levels(df2$ExpYrCat)
    ) %>% mutate(Treatment_Combo = "N"))
  
  # design matrix
  fixed_formula <- reformulate(attr(terms(model), "term.labels"))
  X   <- model.matrix(fixed_formula, newdat)
  beta <- fixef(model)
  vcv  <- vcov(model)
  
  # predictions
  newdat$linpred <- as.vector(X %*% beta)
  newdat$se_linpred <- sqrt(diag(X %*% vcv %*% t(X)))
  newdat$predicted <- exp(newdat$linpred)
  newdat$se_predicted <- newdat$predicted * newdat$se_linpred
  
  # treatment effects
  newdat %>%
    group_by(Treatment_Combo, ExpYrCat) %>%
    summarise(
      control_linpred = linpred[1],
      treated_linpred = linpred[2],
      diff_linpred = treated_linpred - control_linpred,
      ratio = exp(diff_linpred),
      se_diff_linpred = {
        Xg <- X[cur_group_rows(), , drop = FALSE]
        cvec <- Xg[2, ] - Xg[1, ]
        sqrt(as.vector(cvec %*% vcv %*% cvec))
      },
      Measure = measure,
      Phase   = phase,
      .groups = "drop"
    )
}

# now loop through your models + labels
models <- list(
  list(model = yr_effects_start_leaves,   measure = "start", phase = "Leaves"),
  list(model = yr_effects_start_flowers,  measure = "start", phase = "OpenFlowers"),
  list(model = yr_effects_start_fruits,   measure = "start", phase = "RipeFruits"),
  list(model = yr_effects_end_leaves,     measure = "end",   phase = "Leaves"),
  list(model = yr_effects_end_flowers,    measure = "end",   phase = "OpenFlowers"),
  list(model = yr_effects_end_fruits,     measure = "end",   phase = "RipeFruits")
)

# apply function to each and bind results
treatment_effects_all <- purrr::map_dfr(models, ~get_treatment_effects(.x$model, .x$measure, .x$phase))

# PLOT TREATMENT EFFECTS
# Step 0: List the Models & Variables that are Significant in Each Model
    anova(yr_effects_start_leaves) # sig year effect
    anova(yr_effects_start_flowers) # sig year effect
    anova(yr_effects_start_fruits) # sig year effect
  
    anova(yr_effects_end_leaves) # sig year effect
    anova(yr_effects_end_flowers) # sig year effect
    anova(yr_effects_end_fruits) # sig year effect
    
    # Make ANOVA table out of predicts output
    yr_effects_start_leaves_aov <- data.frame(Anova(yr_effects_start_leaves)[3], anova(yr_effects_start_leaves)[4]); yr_effects_start_leaves_aov <- yr_effects_start_leaves_aov %>% mutate(Measure = rep("start"), Phase = rep("Leaves"))
    yr_effects_start_flowers_aov <- data.frame(Anova(yr_effects_start_flowers)[3], anova(yr_effects_start_flowers)[4]); yr_effects_start_flowers_aov <- yr_effects_start_flowers_aov %>% mutate(Measure = rep("start"), Phase = rep("OpenFlowers"))
    yr_effects_start_fruits_aov <- data.frame(Anova(yr_effects_start_fruits)[3], anova(yr_effects_start_fruits)[4]); yr_effects_start_fruits_aov <- yr_effects_start_fruits_aov %>% mutate(Measure = rep("start"), Phase = rep("RipeFruits"))
    
    yr_effects_end_leaves_aov <- data.frame(Anova(yr_effects_end_leaves)[3], anova(yr_effects_end_leaves)[4]); yr_effects_end_leaves_aov <- yr_effects_end_leaves_aov %>% mutate(Measure = rep("end"), Phase = rep("Leaves"))
    yr_effects_end_flowers_aov <- data.frame(Anova(yr_effects_end_flowers)[3], anova(yr_effects_end_flowers)[4]); yr_effects_end_flowers_aov <- yr_effects_end_flowers_aov %>% mutate(Measure = rep("end"), Phase = rep("OpenFlowers"))
    yr_effects_end_fruits_aov <- data.frame(Anova(yr_effects_end_fruits)[3], anova(yr_effects_end_fruits)[4]); yr_effects_end_fruits_aov <- yr_effects_end_fruits_aov %>% mutate(Measure = rep("end"), Phase = rep("RipeFruits"))
    
    # bind
    aov_preds <- bind_rows(yr_effects_start_leaves_aov,
                           yr_effects_start_flowers_aov,
                           yr_effects_start_fruits_aov,
                           yr_effects_end_leaves_aov,
                           yr_effects_end_flowers_aov,
                           yr_effects_end_fruits_aov)
    # add significance codes
    aov_preds <- aov_preds %>% mutate(Significance = ifelse(Pr..Chisq. < 0.001, "***",
                                                            ifelse(Pr..Chisq. < 0.01, "**", 
                                                                   ifelse(Pr..Chisq. < 0.05, "*", 
                                                                          ifelse(Pr..Chisq. < 0.1, ".", ""))))) %>%
      # create row names for join to preds dataframe (below)
      rownames_to_column(var = "rowname") %>%
      mutate(Treatment_Combo = case_when(
        grepl("Temp", rowname) ~ "Warming",
        grepl("Water", rowname) ~ "-W",
        grepl("N", rowname) ~ "+N",
        grepl("CO2", rowname) ~ "+C",
        TRUE ~ ""
      )) %>%
      mutate(Treatment_Combo = paste0(
        ifelse(grepl("Temp", rowname), "+T", ""),
        ifelse(grepl("Water", rowname), "-W", ""),
        ifelse(grepl("N", rowname), "+N", ""),
        ifelse(grepl("CO2", rowname), "+C", "")
      )) %>%
      column_to_rownames(var = "rowname") %>%
      mutate(Treatment_Combo = ifelse(Treatment_Combo == "+N+C", "+C+N", Treatment_Combo)) %>%
      # Add ExpYrCat to rows if applicable
      mutate(
        Treatment_Combo = if_else(
          grepl("ExpYrCat", rownames(.)),  # Check if row name contains "ExpYrCat"
          if_else(
            Treatment_Combo == "",         # If Treatment_Combo is empty
            "ExpYrCat",                    # Just insert "ExpYrCat"
            paste0(Treatment_Combo, " :ExpYrCat")  # Otherwise, append it
          ),
          Treatment_Combo                 # If row name doesn't contain "ExpYrCat", leave as is
        ),
        Treatment_Combo = ifelse(Treatment_Combo == "ExpYrCat", "Ctrl :ExpYrCat", Treatment_Combo)
      )
    
    aov_preds$Treatment_Combo <- gsub("ExpYrCat", " Year", aov_preds$Treatment_Combo)
      
    # check out DFs via AICs
    AICc(yr_effects_start_leaves,
         yr_effects_start_flowers,
         yr_effects_start_fruits,
         yr_effects_end_leaves,
         yr_effects_end_flowers,
         yr_effects_end_fruits)
    
    aov_preds_exp <- aov_preds %>%
      dplyr::select(Measure, Phase, Treatment_Combo, F.value, Pr..Chisq., Significance) %>%
      rename(p.value = Pr..Chisq.)

    # print out model statistics
    print(aov_preds_exp)
    # Unlike linear mixed models (lmer), generalized linear mixed models (glmer) do not directly compute denominator degrees of freedom (DF). This is because glmer uses maximum likelihood (ML) or restricted maximum likelihood (REML) estimation, which does not inherently assign DF to fixed effects.

# Step 1: Generate a New Datasets for Predictions
    # Create a new datasets for predictions
    newdat <- expand.grid(
          Temp.Treatment = c("HTamb", "HTelv"),
          Water.Treatment = c("H2Oamb", "H2Oneg"),
          CO2.Treatment = c("ambient CO2", "+CO2"),
          N.Treatment = c("ambient N", "+N"),
          ExpYrCat = c("1", "2", "3", "4", "5", "6", "7", "8", "10", "11", "12")
        ) %>% # now we can tailor this dataset 
        mutate(
        # add treatment combo column
        Treatment_Combo = case_when(
          Temp.Treatment == "HTelv" & Water.Treatment == "H2Oamb" & CO2.Treatment == "ambient CO2" & N.Treatment == "ambient N" ~ "+T",
          Temp.Treatment == "HTamb" & Water.Treatment == "H2Oneg" & CO2.Treatment == "ambient CO2" & N.Treatment == "ambient N" ~ "-W",
          Temp.Treatment == "HTamb" & Water.Treatment == "H2Oamb" & CO2.Treatment == "+CO2" & N.Treatment == "ambient N" ~ "+C",
          Temp.Treatment == "HTamb" & Water.Treatment == "H2Oamb" & CO2.Treatment == "ambient CO2" & N.Treatment == "+N" ~ "+N",
          Temp.Treatment == "HTamb" & Water.Treatment == "H2Oamb" & CO2.Treatment == "ambient CO2" & N.Treatment == "ambient N" ~ "Ctrl", # All ambient
          TRUE ~ paste0(
            ifelse(Temp.Treatment == "HTelv", "+T", ""),
            ifelse(Water.Treatment == "H2Oneg", "-W", ""),
            ifelse(CO2.Treatment == "+CO2", "+C", ""),
            ifelse(N.Treatment == "+N", "+N", "")
          )
        )
      ) %>%
      # remove instances of 3 & 4-way interactions
      mutate(
        non_ambient_count = rowSums(data.frame(
          Temp.Treatment != "HTamb",
          Water.Treatment != "H2Oamb",
          CO2.Treatment != "ambient CO2",
          N.Treatment != "ambient N"
        ))
      ) %>%
      filter(non_ambient_count <= 2) %>% # Keep only rows with <= 2 non-ambient variables
      dplyr::select(-non_ambient_count) # Drop the helper column
    
# Step 2: Generate specific datasets for models
    # START OF LEAVES
    # Significant: Temp.Treatment, Temp.Treatment:N.Treatment (added N.Treatment by itself too...)
    newdat_yr_effects_start_leaves <- newdat %>%
      # filter to get specific instances we're testing with this response
      filter(Treatment_Combo %in% c("Ctrl", "+T", "-W", "+C", "+N"))
    
    # START OF FLOWERS
    # Significant: Temp.Treatment 
    newdat_yr_effects_start_flowers <- newdat %>%
      # filter to get specific instances we're testing with this response
      filter(Treatment_Combo %in% c("Ctrl", "+T", "-W", "+C", "+N"))
    
    # START OF FRUITS
    # Significant: Temp.Treatment, Water.Treatment
    newdat_yr_effects_start_fruits <- newdat %>%
      # filter to get specific instances we're testing with this response
      filter(Treatment_Combo %in% c("Ctrl", "+T", "-W", "+C", "+N"))
    
    # END OF LEAVES
    # Significant: Temp.Treatment:CO2.Treatment (added Temp.Treatment and CO2.Treatment by themselves too...)
    newdat_yr_effects_end_leaves <- newdat %>%
      # filter to get specific instances we're testing with this response
      filter(Treatment_Combo %in% c("Ctrl", "+T", "-W", "+C", "+N"))# %>% # Formerly:  "Ctrl", "+T", "+T+C", "+C"
      # fix bug
      #mutate(Treatment_Combo = ifelse(Treatment_Combo == "+C+N", "+N+C",Treatment_Combo))
    
    # END OF FLOWERS
    # Significant: Water.Treatment 
    newdat_yr_effects_end_flowers <- newdat %>%
      # filter to get specific instances we're testing with this response
      filter(Treatment_Combo %in% c("Ctrl", "+T", "-W", "+C", "+N")) # formerly "Ctrl", "-W"
    
    # END OF FRUITS
    # Significant: none, but each variable is included separately for kicks
    newdat_yr_effects_end_fruits <- newdat %>%
      # filter to get specific instances we're testing with this response
      filter(Treatment_Combo %in% c("Ctrl", "+T", "-W", "+C", "+N"))  

# Step 3: Extract predictors and standard errors (RUN IN SEQUENCE!!!!)
    # START OF LEAVES
    # Generate Model Matrix for New Data
    fixed_formula <- reformulate(attr(terms(yr_effects_start_leaves), "term.labels"))  # Extract fixed effects formula
    X <- model.matrix(fixed_formula, newdat_yr_effects_start_leaves)  # Generate model matrix
    # Extract Fixed Effect Estimates and Variance-Covariance Matrix
    beta_hat <- fixef(yr_effects_start_leaves)  # Fixed effects
    vcov_beta <- vcov(yr_effects_start_leaves)  # Covariance matrix
    # Compute Predictions and Standard Errors
    link_predictions <- X %*% beta_hat  # Predictions on link scale
    link_se <- sqrt(diag(X %*% vcov_beta %*% t(X)))  # Standard errors
    # Store Predictions in Dataframe
    newdat_start_leaves1 <- newdat_yr_effects_start_leaves %>%
      mutate(
        link_se = exp(link_se),
        predicted = exp(link_predictions),  # Poisson uses log-link, so exp() transforms predictions back
        lower = exp(link_predictions - 1.96 * link_se),  # Confidence interval lower bound
        upper = exp(link_predictions + 1.96 * link_se),  # Confidence interval upper bound
        Measure = "start",  # Label for Measure
        Phase = "Leaves"  # Label for Phase
      )
      # Verify Predictions Using predict()
      newdat_start_leaves1 <- newdat_start_leaves1 %>%
        mutate(
          predict_check = predict(yr_effects_start_leaves, newdata = newdat_yr_effects_start_leaves, type = "response", re.form = NA)
        ) %>%
        dplyr::select(ExpYrCat,Treatment_Combo,Measure,Phase,predicted, link_se) 
      
    # START OF FLOWERS
    # Generate Model Matrix for New Data
    fixed_formula <- reformulate(attr(terms(yr_effects_start_flowers), "term.labels"))  # Extract fixed effects formula
    X <- model.matrix(fixed_formula, newdat_yr_effects_start_flowers)  # Generate model matrix
    # Extract Fixed Effect Estimates and Variance-Covariance Matrix
    beta_hat <- fixef(yr_effects_start_flowers)  # Fixed effects
    vcov_beta <- vcov(yr_effects_start_flowers)  # Covariance matrix
    # Compute Predictions and Standard Errors
    link_predictions <- X %*% beta_hat  # Predictions on link scale
    link_se <- sqrt(diag(X %*% vcov_beta %*% t(X)))  # Standard errors
    # Store Predictions in Dataframe
    newdat_start_flowers1 <- newdat_yr_effects_start_flowers %>%
      mutate(
        link_se = exp(link_se),
        predicted = exp(link_predictions),  # Poisson uses log-link, so exp() transforms predictions back
        lower = exp(link_predictions - 1.96 * link_se),  # Confidence interval lower bound
        upper = exp(link_predictions + 1.96 * link_se),  # Confidence interval upper bound
        Measure = "start",  # Label for Measure
        Phase = "OpenFlowers"  # Label for Phase
      )
      # Verify Predictions Using predict()
      newdat_start_flowers1 <- newdat_start_flowers1 %>%
        mutate(
          predict_check = predict(yr_effects_start_flowers, newdata = newdat_yr_effects_start_flowers, type = "response", re.form = NA)
        )%>%
        dplyr::select(ExpYrCat,Treatment_Combo,Measure,Phase,predicted, link_se) 
        
    # START OF FRUITS
    # Generate Model Matrix for New Data
    fixed_formula <- reformulate(attr(terms(yr_effects_start_fruits), "term.labels"))  # Extract fixed effects formula
    X <- model.matrix(fixed_formula, newdat_yr_effects_start_fruits)  # Generate model matrix
    # Extract Fixed Effect Estimates and Variance-Covariance Matrix
    beta_hat <- fixef(yr_effects_start_fruits)  # Fixed effects
    vcov_beta <- vcov(yr_effects_start_fruits)  # Covariance matrix
    # Compute Predictions and Standard Errors
    link_predictions <- X %*% beta_hat  # Predictions on link scale
    link_se <- sqrt(diag(X %*% vcov_beta %*% t(X)))  # Standard errors
    # Store Predictions in Dataframe
    newdat_start_fruits1 <- newdat_yr_effects_start_fruits %>%
      mutate(
        link_se = exp(link_se),
        predicted = exp(link_predictions),  # Poisson uses log-link, so exp() transforms predictions back
        lower = exp(link_predictions - 1.96 * link_se),  # Confidence interval lower bound
        upper = exp(link_predictions + 1.96 * link_se),  # Confidence interval upper bound
        Measure = "start",  # Label for Measure
        Phase = "RipeFruits"  # Label for Phase
      )
    # Verify Predictions Using predict()
    newdat_start_fruits1 <- newdat_start_fruits1 %>%
      mutate(
        predict_check = predict(yr_effects_start_fruits, newdata = newdat_yr_effects_start_fruits, type = "response", re.form = NA)
      )%>%
        dplyr::select(ExpYrCat,Treatment_Combo,Measure,Phase,predicted, link_se) 
      
    # END OF LEAVES
    # Extract the fixed-effects formula (no response variable, no random effects)
        fixed_formula <- reformulate(attr(terms(yr_effects_end_leaves), "term.labels"))
        # Generate the model matrix for the new data
        X <- model.matrix(fixed_formula, newdat_yr_effects_end_leaves)
        # Extract fixed-effect estimates and covariance matrix
        beta_hat <- fixef(yr_effects_end_leaves)  # Fixed effects
        vcov_beta <- vcov(yr_effects_end_leaves)  # Covariance matrix of fixed effects
        # Calculate predictions on the link scale
        link_predictions <- X %*% beta_hat
        # Standard errors for the link scale predictions
        link_se <- sqrt(diag(X %*% vcov_beta %*% t(X)))
        # Generate predictions
        end_leaves_predictions <- predict(yr_effects_end_leaves, newdata = newdat_yr_effects_end_leaves, type = "link", re.form = NA) 
        # Back-transform predictions and confidence intervals
        newdat_end_leaves1 <- newdat_yr_effects_end_leaves %>%
          mutate(
                link_se = (exp(link_se))^2,
                predicted = 365 - (exp(link_predictions))^2 - 1,  # Back-transform predictions
                # lower = (exp(link_predictions - 1.96 * link_se))^2,  # Lower bound
                # upper = (exp(link_predictions + 1.96 * link_se))^2,  # Upper bound

            # Back-transform predictions to the original response scale
            # predicted = 365 - (link_predictions^2 - 1),
            # Back-transform link_se to the response scale
            #link_se <- exp(link_predictions) * link_se,
    
            Measure = "start",  # Label for Measure
            Phase = "Leaves"  # Label for Phase
          ) %>% 
          mutate(Measure = rep("end"), Phase = rep("Leaves")) #%>% dplyr::select(-c(se_sqrt,predicted_sqrt,lower_sqrt,upper_sqrt))
        # Verify Predictions Using predict()
        newdat_end_leaves1 <- newdat_end_leaves1 %>%
          mutate(
            predict_check = predict(yr_effects_end_leaves, newdata = newdat_yr_effects_end_leaves, type = "response", re.form = NA)
          )%>%
        dplyr::select(ExpYrCat,Treatment_Combo,Measure,Phase,predicted, link_se) 
        
    # END OF FLOWERS
    # Extract the fixed-effects formula (no response variable, no random effects)
        fixed_formula <- reformulate(attr(terms(yr_effects_end_flowers), "term.labels"))
        # Generate the model matrix for the new data
        X <- model.matrix(fixed_formula, newdat_yr_effects_end_flowers)
        # Extract fixed-effect estimates and covariance matrix
        beta_hat <- fixef(yr_effects_end_flowers)  # Fixed effects
        vcov_beta <- vcov(yr_effects_end_flowers)  # Covariance matrix of fixed effects
        # Calculate predictions on the link scale
        link_predictions <- X %*% beta_hat
        # Standard errors for the link scale predictions
        link_se <- sqrt(diag(X %*% vcov_beta %*% t(X)))
        # Generate predictions
        end_flowers_predictions <- predict(yr_effects_end_flowers, newdata = newdat_yr_effects_end_flowers, type = "link", re.form = NA) 
        # Back-transform predictions and confidence intervals
        newdat_end_flowers1 <- newdat_yr_effects_end_flowers %>%
          mutate(
                link_se = (exp(link_se))^2,
                predicted = 365 - (exp(link_predictions))^2 - 1,  # Back-transform predictions
                # lower = (exp(link_predictions - 1.96 * link_se))^2,  # Lower bound
                # upper = (exp(link_predictions + 1.96 * link_se))^2,  # Upper bound

            # Back-transform predictions to the original response scale
            # predicted = 365 - (link_predictions^2 - 1),
            # Back-transform link_se to the response scale
            #link_se <- exp(link_predictions) * link_se,
    
            Measure = "start",  # Label for Measure
            Phase = "Leaves"  # Label for Phase
          ) %>% 
          mutate(Measure = rep("end"), Phase = rep("OpenFlowers")) #%>% dplyr::select(-c(se_sqrt,predicted_sqrt,lower_sqrt,upper_sqrt))
        # Verify Predictions Using predict()
        newdat_end_flowers1 <- newdat_end_flowers1 %>%
          mutate(
            predict_check = predict(yr_effects_end_flowers, newdata = newdat_yr_effects_end_flowers, type = "response", re.form = NA)
          )%>%
        dplyr::select(ExpYrCat,Treatment_Combo,Measure,Phase,predicted, link_se) 
        
    # END OF FRUITS
    # Extract the fixed-effects formula (no response variable, no random effects)
        fixed_formula <- reformulate(attr(terms(yr_effects_end_fruits), "term.labels"))
        # Generate the model matrix for the new data
        X <- model.matrix(fixed_formula, newdat_yr_effects_end_fruits)
        # Extract fixed-effect estimates and covariance matrix
        beta_hat <- fixef(yr_effects_end_fruits)  # Fixed effects
        vcov_beta <- vcov(yr_effects_end_fruits)  # Covariance matrix of fixed effects
        # Calculate predictions on the link scale
        link_predictions <- X %*% beta_hat
        # Standard errors for the link scale predictions
        link_se <- sqrt(diag(X %*% vcov_beta %*% t(X)))
        # Generate predictions
        end_fruits_predictions <- predict(yr_effects_end_fruits, newdata = newdat_yr_effects_end_fruits, type = "link", re.form = NA) 
        # Back-transform predictions and confidence intervals
        newdat_end_fruits1 <- newdat_yr_effects_end_fruits %>%
          mutate(
                link_se = (exp(link_se))^2,
                predicted = 365 - (exp(link_predictions))^2 - 1,  # Back-transform predictions
            Measure = "start",  # Label for Measure
            Phase = "Leaves"  # Label for Phase
          ) %>% 
          mutate(Measure = rep("end"), Phase = rep("RipeFruits")) #%>% dplyr::select(-c(se_sqrt,predicted_sqrt,lower_sqrt,upper_sqrt))
        # Verify Predictions Using predict()
        newdat_end_fruits1 <- newdat_end_fruits1 %>%
          mutate(
            predict_check = predict(yr_effects_end_fruits, newdata = newdat_yr_effects_end_fruits, type = "response", re.form = NA)
          )%>%
        dplyr::select(ExpYrCat,Treatment_Combo,Measure,Phase,predicted, link_se) 
        
        # Combine dataframes
        newdata_all_models <- bind_rows(newdat_start_leaves1, 
                                        newdat_start_flowers1, 
                                        newdat_start_fruits1, 
                                        newdat_end_leaves1, 
                                        newdat_end_flowers1, 
                                        newdat_end_fruits1)
        # Add anova significance codes
        newdata_all_models <- newdata_all_models %>%
          # specify that these are all instances where ExpYr interacts with treatment (technically)
          mutate(Treatment_Combo = ifelse(Treatment_Combo == "Ctrl", "Ctrl : Year", Treatment_Combo)) %>%
          #mutate(Treatment_Combo = paste(Treatment_Combo, ":ExpYrCat")) %>%
          left_join(., aov_preds, by = c("Treatment_Combo", "Phase", "Measure")) 
        
# create treatment effects dataframe
treatment_effects_all <- newdata_all_models %>%
  group_by(ExpYrCat, Phase, Measure) %>%
  # Separate Ctrl rows
  mutate(is_ctrl = Treatment_Combo == "Ctrl : Year") %>%
  # Create separate dataframes for Ctrl and non-Ctrl
  group_split() %>%
  map_df(function(df_group) {
    ctrl_row <- df_group %>% filter(is_ctrl)
    non_ctrl_rows <- df_group %>% filter(!is_ctrl)

    # Join Ctrl values to each non-Ctrl row
    non_ctrl_rows %>%
      left_join(ctrl_row %>% 
                  dplyr::select(predicted_ctrl = predicted, 
                         link_se_ctrl = link_se,
                         ExpYrCat, Phase, Measure),
                by = c("ExpYrCat", "Phase", "Measure")) %>%
      mutate(
        predicted_diff = predicted - predicted_ctrl,
        link_se_diff = sqrt(link_se^2 + link_se_ctrl^2)
      )
  }) %>%
  dplyr::select(ExpYrCat, Phase, Measure, Treatment_Combo, predicted_diff, link_se_diff)
        
    
# Step 4: Plots
        treatment_effects_all$Measure <- factor(treatment_effects_all$Measure, levels = c("start", "end"))
        newdata_all_models$Measure <- factor(newdata_all_models$Measure, levels = c("start", "end"))
        # labels for plots
        Phase.labs <- c("Leaves", "Open Flowers", "Ripe Fruits")
        names(Phase.labs) <- c("Leaves", "OpenFlowers", "RipeFruits")
         
        Measure.labs <- c("Start", "End")
        names(Measure.labs) <- c("start", "end")

newdata_all_models$Treatment_Combo <- factor(newdata_all_models$Treatment_Combo, levels = c("Ctrl :ExpYrCat", "+T :ExpYrCat", "-W :ExpYrCat", "+C :ExpYrCat", "+N :ExpYrCat"))
treatment_effects_all$Treatment_Combo <- factor(treatment_effects_all$Treatment_Combo, levels = c("+T", "-W", "+C", "+N"))

# WITH treatment effects
treatment_effects_all_2 <- treatment_effects_all %>%
  mutate(ExpYrNum = as.numeric(ifelse(ExpYrCat == 1,"2012", 
                           ifelse(ExpYrCat ==  2, "2013",
                                  ifelse(ExpYrCat ==  3, "2014",
                                         ifelse(ExpYrCat == 4, "2015", 
                                                ifelse(ExpYrCat ==  5, "2016",
                                                       ifelse(ExpYrCat == 6, "2017",
                                                              ifelse(ExpYrCat == 7, "2018",
                                                                     ifelse(ExpYrCat == 8,"2019",
                                                                            ifelse(ExpYrCat ==  9, "2020",
                                                                                   ifelse(ExpYrCat == 10, "2021",
                                                                                          ifelse(ExpYrCat == 11, "2022",
                                                                                                 ifelse(ExpYrCat == 12, 2023, NA)))))))))))))) #%>% 

plot_data <- treatment_effects_all_2 %>%
  left_join(cumulative_gdd_year, by = c("ExpYrNum" = "Year")) 

plot_data %>%
  mutate(PhaseMeasure = paste(Measure, Phase, sep = " of "),
         PhaseMeasure = ifelse(PhaseMeasure == "end of Leaves", "End of Leaves",
                               ifelse(PhaseMeasure == "start of Leaves", "Start of Leaves",
                                      ifelse(PhaseMeasure == "end of OpenFlowers", "End of Flowers",
                                             ifelse(PhaseMeasure == "start of OpenFlowers", "Start of Flowers",
                                                    ifelse(PhaseMeasure == "start of RipeFruits", "Start of Fruits", 
                                                           ifelse(PhaseMeasure == "end of RipeFruits", "End of Fruits", PhaseMeasure)))))),
         PhaseMeasure = factor(PhaseMeasure, levels = c("Start of Leaves", "End of Leaves", "Start of Flowers", "End of Flowers", "Start of Fruits", "End of Fruits"))) %>%
ggplot(. , aes(x = predicted_diff, y = as.factor(ExpYrNum), color = Treatment_Combo)) +
  geom_point(position = position_dodge(0.5), size = 1.75) +
  geom_vline(xintercept = 0, col = "black", linetype = "dashed") +
  geom_errorbar(aes(xmin = predicted_diff - link_se_diff, xmax = predicted_diff + link_se_diff), position = position_dodge(0.5), width = 0) +
  labs(
   # title = "Model Predictions for Phase Start and End by Year",
    x = "Treatment response (Mean Day of Year ± 1 SE)",
    y = "Year of the experiment"
  ) +
  theme(
    panel.border = element_rect(color = "black", fill = NA),
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = "grey90", size = 0.25),
    panel.grid.minor = element_blank(), 
    strip.background.x=element_rect(color = NA,  fill=NA), 
    strip.background.y=element_rect(color = "black",  fill=NA),
    strip.text = element_text(size = 10, face = "bold"),
    plot.title = element_text(hjust = 0.5),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 12), #face = "bold"
    legend.position = "left",
    legend.background = element_rect(fill = "white"),
    legend.title = element_text()
  ) +
  scale_color_manual(values = c("+T"="firebrick1",
                                "-W"="cornflowerblue",
                                "+C"="grey50",
                                "+N"="palegreen3"), name = "Treatment") +
  facet_wrap(. ~ PhaseMeasure, ncol = 2
             )
# interannual variability section
```
[Back to Index](#index)
 
################################################################################
# Select, run and graph models
################################################################################
# STEPS:
# 1) run iterative models to extract least-complicated-order model based on p-values
# 2) print out best-fit/final models for each response combination
# 3) run predict() sequence for that model combination

When to Use lmer() vs. glmer() models (Decision Guide)
Criterion                   |	lmer()	                    | glmer()
----------------------------|-----------------------------|---------------------------
Response Variable Type      |	Continuous (e.g., time)	    | Count, binary, proportions
Residual Distribution	      | Normal	                    | Non-normal, skewed
Overdispersion Detected?    |	No	                        | Yes (variance > mean)
Interpretability Needed?    |	Easier	                    | More complex but flexible

Notes on distributions
Distribution                    | Use Case	                      | glmer() Syntax
--------------------------------|------------------------.--------|-----------------------------
Poisson	                        | Count data, no overdispersion   | family = poisson
Negative Binomial	              | Overdispersed count data        | glmer.nb()
Gamma	                          | Continuous, right-skewed data   |	family = Gamma(link = "log")
Zero-Inflated Poisson	          | Excess zeros in count data      | glmmTMB() (family = poisson)
Zero-Inflated Negative Binomial	| Excess zeros + overdispersion   | glmmTMB() (family = nbinom2)

 # *Some notes on model distribution and selection with AICc:*
     # If AICc < 2 units difference → Choose the simpler model for both datasets.
     # If AICc > 2 units difference → Stick with the model that fits each dataset better individually.
     # Count Data Consistency: If both datasets are count-based, Poisson or Negative Binomial is preferred.
     
# Decision Hierarchy for My Data (Best Practices):
    First: Check if overdispersion is present (I've done this already in an earlier chunk).
    Then: If overdispersion is not severe, use a Poisson GLMM.
    If overdispersion is present: use continuous (Non-negative) Gamma GLMM (glmer)
    For Excess Zeros: Try Zero-Inflated Models as a last resort.
  
# If overdispersion is present (in the case of non-transformed variables), prioritize handling it (Negative Binomial or Gamma).
# If both count data + overdispersion, Use a Square-Root transformation on the response varible

###################### some notes on back-transformations ######################

    Back-transforming is necessary when the response variable has been transformed (e.g., square-root, log, or Box-Cox transformation) to meet model assumptions. The goal is to return predicted values and confidence intervals to the original scale for meaningful interpretation.

    *Why We Do It:*
    - Interpretability: Model coefficients and predictions are in the transformed space, which can be misleading if not reversed.
    - Comparability: Allows direct comparison with raw data or other studies using untransformed scales.
    - Correct Error Propagation: Ensures confidence intervals and effect sizes reflect the original measurement scale.
    *When We Do It:*
    - After obtaining predictions (fitted values) from the model.
    - Before visualizing or reporting results.
    - When presenting effect sizes, estimated means, or confidence intervals.
    
    - Some of the models below use a Gamma (log-link) distribution, predicted values are exponentiated. Thus, we need to undo both transformations:

  *Back-transformed value = (e^predicted)^2*

#### Model Selection: selection for best-fit models, given interacting treatment effects {#model_selection}
    Why GLMM over LMM?
    + Count data is non-Gaussian and right-skewed, violating assumptions of LMM.
    + Poisson/Negative Binomial GLMM can model the discrete nature of data more appropriately.
    + GLMMs allow for non-normal distributions and address overdispersion more effectively.
    
    The following random effects were included in the model to account for hierarchical structure:
    (1 | ExpYrCat) + (1 | Species.observed) + (1|Ring/Plot) + (1|ExpYrCat:Species.observed) + (1|Ring:Plot:Species.observed)
    + Plot is nested within Ring.
    + Species identity appears across multiple plots, not exclusively within plots.
    + Including ExpYrCat accounts for temporal variations across years.
    + Variance decomposition (VarCorr()) confirmed low variance for some random effects, suggesting possible overfitting.
```{r}
# ensure temp and water treatments are factors before we start
    df2$Temp.Treatment <- as.factor(df2$Temp.Treatment)
    df2$Water.Treatment <- as.factor(df2$Water.Treatment)
    
# Start with higher-order interactions. If overdispersion is detected, switch from poisson to gamma, compare using AICc
    
# If there is a significant main effect, keep adjascent two-way interactions in final model. 
    
##### Start of leaves ##########################################################
            # 4-way (full-interactive model)
            start_leaves_glmer_poisson_4 <- glmer(start ~ Temp.Treatment * Water.Treatment * CO2.Treatment * N.Treatment + (1 | ExpYrCat) + (1 | Species.observed) + (1|Ring:Plot:Species.observed), data = subset(df2, Phase == "Leaves"), family = poisson)
            check_overdispersion(start_leaves_glmer_poisson_4) # good
            car::vif(start_leaves_glmer_poisson_4) # not good ( should be 4 or less)
            anova(start_leaves_glmer_poisson_4) 
            Anova(start_leaves_glmer_poisson_4)
            # no 3 or 4-way interactions, so moving down to 2 
            
            start_leaves_glmer_poisson_2 <- glmer(start ~ Temp.Treatment + Water.Treatment + CO2.Treatment + N.Treatment + Temp.Treatment:Water.Treatment + Temp.Treatment:CO2.Treatment + Temp.Treatment:N.Treatment + Water.Treatment:CO2.Treatment + Water.Treatment:N.Treatment + CO2.Treatment:N.Treatment + (1 | ExpYrCat) + (1 | Species.observed) + (1|Ring:Plot:Species.observed), data = subset(df2, Phase == "Leaves"), family = poisson)
            check_overdispersion(start_leaves_glmer_poisson_2)
            car::vif(start_leaves_glmer_poisson_2) # better
            anova(start_leaves_glmer_poisson_2)
            Anova(start_leaves_glmer_poisson_2)
        # Significant: Temp.Treatment, Temp.Treatment:N.Treatment

        # glmer for predict():
        # keep Temp.Treatment-adjacent interactions
        start_leaves_preds <- glmer(start ~ Temp.Treatment + Water.Treatment + N.Treatment + CO2.Treatment + Temp.Treatment:Water.Treatment + Temp.Treatment:N.Treatment + Temp.Treatment:CO2.Treatment + Water.Treatment:N.Treatment + Water.Treatment:CO2.Treatment + N.Treatment:CO2.Treatment + (1 | ExpYrCat) + (1 | Species.observed) + (1|Ring:Plot:Species.observed), data = subset(df2, Phase == "Leaves"), family = poisson)


##### Start of flowers #########################################################
            # start with higher-order interactions
            # 4-way (full-interactive model)
            start_flowers_glmer_poisson_4 <- glmer(start ~ Temp.Treatment * Water.Treatment * CO2.Treatment * N.Treatment + (1 | ExpYrCat) + (1 | Species.observed) + (1|Ring:Plot:Species.observed), data = subset(df2, Phase == "OpenFlowers"), family = poisson)
            check_overdispersion(start_flowers_glmer_poisson_4) 
            car::vif(start_flowers_glmer_poisson_4) # not good
            anova(start_flowers_glmer_poisson_4) 
            Anova(start_flowers_glmer_poisson_4)
            # no 3 or 4-way interactions, so moving down to 2
            
            start_flowers_glmer_poisson_2 <- glmer(start ~ Temp.Treatment + Water.Treatment + CO2.Treatment + N.Treatment + Temp.Treatment:Water.Treatment + Temp.Treatment:CO2.Treatment + Temp.Treatment:N.Treatment + Water.Treatment:CO2.Treatment + Water.Treatment:N.Treatment + CO2.Treatment:N.Treatment + (1 | ExpYrCat) + (1 | Species.observed) + (1|Ring:Plot:Species.observed), data = subset(df2, Phase == "OpenFlowers"), family = poisson)
            check_overdispersion(start_flowers_glmer_poisson_2)
            car::vif(start_flowers_glmer_poisson_2) 
            anova(start_flowers_glmer_poisson_2)
            Anova(start_flowers_glmer_poisson_2)
        # Significant: Temp.Treatment 
            
        # glmer for predict():
        # keep Temp.Treatment-adjacent interactions
        start_flowers_preds <- glmer(start ~ Temp.Treatment + Water.Treatment + N.Treatment + CO2.Treatment + Temp.Treatment:Water.Treatment + Temp.Treatment:N.Treatment + Temp.Treatment:CO2.Treatment + Water.Treatment:N.Treatment + Water.Treatment:CO2.Treatment + N.Treatment:CO2.Treatment + (1 | ExpYrCat) + (1 | Species.observed) + (1|Ring:Plot:Species.observed), data = subset(df2, Phase == "OpenFlowers"), family = poisson)
        # OR:
        start_flowers_preds <- glmer(start ~ Temp.Treatment + Water.Treatment + N.Treatment + CO2.Treatment + (1 | ExpYrCat) + (1 | Species.observed) + (1|Ring:Plot:Species.observed), data = subset(df2, Phase == "OpenFlowers"), family = poisson)
            
    
##### Start of fruits ##########################################################
            # start with higher-order interactions
            # 4-way (full-interactive model)
            start_fruits_glmer_poisson_4 <- glmer(start ~ Temp.Treatment * Water.Treatment * CO2.Treatment * N.Treatment + (1 | ExpYrCat) + (1 | Species.observed) + (1|Ring:Plot:Species.observed), data = subset(df2, Phase == "RipeFruits"), family = poisson)
            check_overdispersion(start_fruits_glmer_poisson_4)
            car::vif(start_fruits_glmer_poisson_4) # not good
            anova(start_fruits_glmer_poisson_4) 
            Anova(start_fruits_glmer_poisson_4)
            # no 3 or 4-way interactions, so moving down to 2
            
            start_fruits_glmer_poisson_2 <- glmer(start ~ Temp.Treatment + Water.Treatment + CO2.Treatment + N.Treatment + Temp.Treatment:Water.Treatment + Temp.Treatment:CO2.Treatment + Temp.Treatment:N.Treatment + Water.Treatment:CO2.Treatment + Water.Treatment:N.Treatment + CO2.Treatment:N.Treatment + (1 | ExpYrCat) + (1 | Species.observed) + (1|Ring:Plot:Species.observed), data = subset(df2, Phase == "RipeFruits"), family = poisson)
            check_overdispersion(start_fruits_glmer_poisson_2)
            car::vif(start_fruits_glmer_poisson_2)
            anova(start_fruits_glmer_poisson_2)
            Anova(start_fruits_glmer_poisson_2)
        # Significant: Temp.Treatment, Water.Treatment
            
        # glmer for predict():
        # keep Water-Treatment- and Temp.Treatment-adjacent interactions
        start_fruits_preds <- glmer(start ~ Temp.Treatment + Water.Treatment + Temp.Treatment:Water.Treatment + Temp.Treatment:CO2.Treatment + Temp.Treatment:N.Treatment + Water.Treatment:CO2.Treatment + Water.Treatment:N.Treatment + (1 | ExpYrCat) + (1 | Species.observed) + (1|Ring:Plot:Species.observed), data = subset(df2, Phase == "RipeFruits"), family = poisson)

##### End of leaves ############################################################
            end_leaves_glmer_poisson_4 <- glmer(end_rev ~ Temp.Treatment * Water.Treatment * CO2.Treatment * N.Treatment + (1 | ExpYrCat) + (1 | Species.observed) + (1|Ring:Plot:Species.observed), data = subset(df2, Phase == "Leaves"), family = poisson)
            check_overdispersion(end_leaves_glmer_poisson_4) 
            # overdispersion detected, so transform end variable & use gamma log-link 
            end_leaves_glmer_gamma_2 <- glmer(sqrt_end_rev ~ Temp.Treatment + Water.Treatment + CO2.Treatment + N.Treatment + Temp.Treatment:Water.Treatment + Temp.Treatment:CO2.Treatment + Temp.Treatment:N.Treatment + Water.Treatment:CO2.Treatment + Water.Treatment:N.Treatment + CO2.Treatment:N.Treatment + (1 | ExpYrCat) + (1 | Species.observed) + (1|Ring:Plot:Species.observed), data = subset(df2, Phase == "Leaves"), family = Gamma(link = "log"))
            # we can't test for overdispersion b/c overdispersion checks can only be used for models from Poisson families or binomial families with trials > 1.
            car::vif(end_leaves_glmer_gamma_2)
            anova(end_leaves_glmer_gamma_2)
            Anova(end_leaves_glmer_gamma_2)
            
            # retroactively check if higher-order interactions are important
            # 4-way (full-interactive model)
            end_leaves_glmer_gamma_4 <- glmer(sqrt_end_rev ~ Temp.Treatment * Water.Treatment * CO2.Treatment * N.Treatment + (1 | ExpYrCat) + (1 | Species.observed) + (1|Ring:Plot:Species.observed), data = subset(df2, Phase == "Leaves"), family = Gamma(link = "log")) 

            car::vif(end_leaves_glmer_gamma_4) # nope
            anova(end_leaves_glmer_gamma_4) 
            Anova(end_leaves_glmer_gamma_4)
            # no 3 or 4-way interactions, so 2-way models (above) was the right choice
            # Significant: Temp.Treatment:CO2.Treatment
            
        # glmer for predict():
        # No main effects here (which is so interesting!), so no need to add any adjacent interactions
        end_leaves_preds <- glmer(sqrt_end_rev ~  Temp.Treatment + Water.Treatment + CO2.Treatment + N.Treatment + Temp.Treatment:Water.Treatment + Temp.Treatment:CO2.Treatment + Temp.Treatment:N.Treatment + Water.Treatment:CO2.Treatment + Water.Treatment:N.Treatment + CO2.Treatment:N.Treatment + (1 | ExpYrCat) + (1 | Species.observed) + (1|Ring:Plot:Species.observed), data = subset(df2, Phase == "Leaves"), family = Gamma(link = "log")) 
        # took out: Temp.Treatment + CO2.Treatment + Temp.Treatment:CO2.Treatment
        # ^^ 2023 update :'(

##### End of flowers ###########################################################
            end_flowers_glmer_poisson_2 <- glmer(end_rev ~ Temp.Treatment + Water.Treatment + CO2.Treatment + N.Treatment + Temp.Treatment:Water.Treatment + Temp.Treatment:CO2.Treatment + Temp.Treatment:N.Treatment + Water.Treatment:CO2.Treatment + Water.Treatment:N.Treatment + CO2.Treatment:N.Treatment + (1 | ExpYrCat) + (1 | Species.observed) + (1|Ring:Plot:Species.observed), data = subset(df2, Phase == "OpenFlowers"), family = poisson)
            check_overdispersion(end_flowers_glmer_poisson_2) 
            # overdispersion detected, so transform end variable
            end_flowers_glmer_gamma_2 <- glmer(sqrt_end_rev ~ Temp.Treatment + Water.Treatment + CO2.Treatment + N.Treatment + Temp.Treatment:Water.Treatment + Temp.Treatment:CO2.Treatment + Temp.Treatment:N.Treatment + Water.Treatment:CO2.Treatment + Water.Treatment:N.Treatment + CO2.Treatment:N.Treatment + (1 | ExpYrCat) + (1 | Species.observed) + (1|Ring:Plot:Species.observed), data = subset(df2, Phase == "OpenFlowers"), family = Gamma(link = "log"))
            car::vif(end_flowers_glmer_gamma_2)
            anova(end_flowers_glmer_gamma_2)
            Anova(end_flowers_glmer_gamma_2)
            
            # retroactively check if higher-order interactions are important
            # 4-way (full-interactive model)
            end_flowers_glmer_gamma_4 <- glmer(sqrt_end_rev ~ Temp.Treatment * Water.Treatment * CO2.Treatment * N.Treatment + (1 | ExpYrCat) + (1 | Species.observed) + (1|Ring:Plot:Species.observed), data = subset(df2, Phase == "OpenFlowers"), family = Gamma(link = "log"))
            # (still) can't check for overdispersion 
            car::vif(end_flowers_glmer_gamma_4) # nope
            anova(end_flowers_glmer_gamma_4) 
            Anova(end_flowers_glmer_gamma_4)
            # no 3 or 4-way interactions, so 2-way models (above) was the right choice!
            
        # Significant: Water.Treatment 
        # nothing significant...put all main effects in model instead
        # glmer for predict():
        # No main effects here (which is so interesting!), so no need to add any adjacent interactions
       end_flowers_preds <- glmer(sqrt_end_rev ~ Temp.Treatment + Water.Treatment + CO2.Treatment + N.Treatment + (1 | ExpYrCat) + (1 | Species.observed) + (1|Ring:Plot:Species.observed), data = subset(df2, Phase == "OpenFlowers"), family = Gamma(link = "log"))
        
##### End of fruits ############################################################
            end_fruits_glmer_poisson_2 <- glmer(end_rev ~ Temp.Treatment + Water.Treatment + CO2.Treatment + N.Treatment + Temp.Treatment:Water.Treatment + Temp.Treatment:CO2.Treatment + Temp.Treatment:N.Treatment + Water.Treatment:CO2.Treatment + Water.Treatment:N.Treatment + CO2.Treatment:N.Treatment + (1 | ExpYrCat) + (1 | Species.observed) + (1|Ring:Plot:Species.observed), data = subset(df2, Phase == "RipeFruits"), family = poisson)
            check_overdispersion(end_fruits_glmer_poisson_2)
            # overdispersion detected, so transform end variable
            end_fruits_glmer_gamma_2 <- glmer(sqrt_end_rev ~ Temp.Treatment + Water.Treatment + CO2.Treatment + N.Treatment + Temp.Treatment:Water.Treatment + Temp.Treatment:CO2.Treatment + Temp.Treatment:N.Treatment + Water.Treatment:CO2.Treatment + Water.Treatment:N.Treatment + CO2.Treatment:N.Treatment + (1 | ExpYrCat) + (1 | Species.observed) + (1|Ring:Plot:Species.observed), data = subset(df2, Phase == "RipeFruits"), family = Gamma(link = "log")) 
            car::vif(end_fruits_glmer_gamma_2)
            anova(end_fruits_glmer_gamma_2)
            Anova(end_fruits_glmer_gamma_2)
            
            # retroactively check if higher-order interactions are important
            # 4-way (full-interactive model)
            end_fruits_glmer_gamma_4 <- glmer(sqrt_end_rev ~ Temp.Treatment * Water.Treatment * CO2.Treatment * N.Treatment + (1 | ExpYrCat) + (1 | Species.observed) + (1|Ring:Plot:Species.observed), data = subset(df2, Phase == "RipeFruits"), family = Gamma(link = "log"))
            # (still) can't check for overdispersion (haha)
            car::vif(end_fruits_glmer_gamma_4) # not good
            anova(end_fruits_glmer_gamma_4) 
            Anova(end_fruits_glmer_gamma_4) # Temp.Treatment:CO2.Treatment:N.Treatment is M.S., but not by much
            # no 3 or 4-way interactions, so 2-way models (above) was the right choice
            
        # Significant: Temp.Treatment:CO2.Treatment
        # glmer for predict():
       end_fruits_preds <- glmer(sqrt_end_rev ~ Temp.Treatment + Water.Treatment + CO2.Treatment + N.Treatment + (1 | ExpYrCat) + (1 | Species.observed) + (1|ExpYrCat:Ring:Species.observed), data = subset(df2, Phase == "RipeFruits"),  family = Gamma(link = "log")) 
```
[Back to Index](#index)

#### graph residuals, address model complexity and overfitting {#model_checks}
```{r}
    par(mfrow = c(2,3))
    plot(start_leaves_glmer_poisson_2, main = "Residuals: Start/Leaves, Poisson")
    plot(start_flowers_glmer_poisson_2, main = "Residuals: Start/Flowers, Poisson")
    plot(start_fruits_glmer_poisson_2, main = "Residuals: Start/Fruits, Poisson")
    
    plot(end_leaves_glmer_gamma_2, main = "Residuals: End/Leaves, Gamma")
    plot(end_flowers_glmer_gamma_2, main = "Residuals: End/Flowers, Gamma")
    plot(end_fruits_glmer_gamma_2, main = "Residuals: End/Fruits, Gamma")
    par(mfrow = c(1,1))

# Run AICc for to get approximate DF for each instance
AICc(start_leaves_glmer_poisson_2,
     start_flowers_glmer_poisson_2,
     start_fruits_glmer_poisson_2,
     
     end_leaves_glmer_gamma_2,
     end_flowers_glmer_gamma_2,
     end_fruits_glmer_gamma_2)
# These df are a little low, but in the ballpark ranges for these variables...

# Multicollinearity Check (VIF):
vif(start_leaves_glmer_poisson_2) # it is similar for other models...
vif(start_flowers_glmer_poisson_2)
vif(start_fruits_glmer_poisson_2)

vif(end_leaves_glmer_gamma_2)
vif(end_flowers_glmer_gamma_2)
vif(end_fruits_glmer_gamma_2)
# No multicollinearity detected (all VIF < 5)

# Cross-Validation Results:
train_control <- trainControl(method = "cv", number = 5)
cv_model <- train(start ~ Temp.Treatment + Water.Treatment + CO2.Treatment + 
                    N.Treatment, data = subset(df2, Phase == "Leaves"), 
                    method = "glm", family = poisson, trControl = train_control)
print(cv_model)
# Result: No substantial model instability detected across folds.

# end residuals, compexity, overfitting
```
[Back to Index](#index)

This workflow (above) rigorously confirms that the GLMM approach with Poisson and Gamma distributions is the most suitable for analyzing our phenological data. This methodology properly handles count data, non-normal distributions, and overdispersion while balancing complexity and model fit.

    A note about getting DF from glmers: this is basically impossible (there is no method for it)
        + GLMMs use maximum likelihood estimation (MLE), which doesn't directly compute degrees of freedom for fixed effects.
        + The asymptotic approximation assumes infinite degrees of freedom for large datasets, which explains the Inf values in our AICc output.
        + Denominator degrees of freedom in mixed models are complex due to the partial pooling of information from random effects.
        + From Ben Bolker: Run the model in lme (if possible) and use the denominator df reported there (which follow a simple ‘inner-outer’ rule which should correspond to the canonical answer for simple/orthogonal designs), applied to t or F tests

#### graph model means and SEs using predict() {#main_graphs}
```{r}
# Step 1: get NumDF and DenDF for all models
    # re-run using nlme to approximate df
    start_leaves_preds_lme <- lme(
      fixed = start ~ Temp.Treatment + Water.Treatment + N.Treatment + CO2.Treatment +
               Temp.Treatment:Water.Treatment + Temp.Treatment:N.Treatment + Temp.Treatment:CO2.Treatment +
               Water.Treatment:N.Treatment + Water.Treatment:CO2.Treatment +
               N.Treatment:CO2.Treatment,
      random = list(ExpYrCat = ~1, Species.observed = ~1, Ring = ~1 | Plot),
      data = subset(df2, Phase == "Leaves")
    );anova(start_leaves_preds_lme)
    
    start_flowers_preds_lme <- lme(
      fixed = start ~ Temp.Treatment,
      random = list(ExpYrCat = ~1, Species.observed = ~1, Ring = ~1 | Plot),
      data = subset(df2, Phase == "OpenFlowers")
    );anova(start_flowers_preds_lme)
    
    start_fruits_preds_lme <- lme(
      fixed = start ~ Temp.Treatment + Water.Treatment + Temp.Treatment:Water.Treatment,
      random = list(ExpYrCat = ~1, Species.observed = ~1, Ring = ~1 | Plot),
      data = subset(df2, Phase == "RipeFruits")
    );anova(start_fruits_preds_lme)
    
    end_leaves_preds_lme <- lme(
      fixed = sqrt_end_rev ~ Temp.Treatment + Water.Treatment + N.Treatment + CO2.Treatment +
               Temp.Treatment:Water.Treatment + Temp.Treatment:N.Treatment + Temp.Treatment:CO2.Treatment +
               Water.Treatment:N.Treatment + Water.Treatment:CO2.Treatment +
               N.Treatment:CO2.Treatment,
      random = list(ExpYrCat = ~1, Species.observed = ~1, Ring = ~1 | Plot),
      data = subset(df2, Phase == "Leaves")
    );anova(end_leaves_preds_lme)
    
    end_flowers_preds_lme <- lme(
      fixed = sqrt_end_rev ~ Temp.Treatment + Water.Treatment + N.Treatment + CO2.Treatment,
      random = list(ExpYrCat = ~1, Species.observed = ~1, Ring = ~1 | Plot),
      data = subset(df2, Phase == "OpenFlowers")
    );anova(end_flowers_preds_lme)
    
    end_fruits_preds_lme <- lme(
      fixed = sqrt_end_rev ~ Temp.Treatment + Water.Treatment + N.Treatment + CO2.Treatment,
      random = list(ExpYrCat = ~1, Species.observed = ~1, Ring = ~1 | Plot),
      data = subset(df2, Phase == "RipeFruits")
    );anova(end_fruits_preds_lme)
    

# Step 2: List the Models & Variables that are Significant in Each Model
    anova(start_leaves_preds) # Significant: Temp.Treatment, Temp.Treatment:N.Treatment (added N.Treatment by itself too...)
    anova(start_flowers_preds) # Significant: Temp.Treatment 
    anova(start_fruits_preds) # Significant: Temp.Treatment, Water.Treatment
  
    anova(end_leaves_preds) # Significant: Temp.Treatment:CO2.Treatment & CO2.Treatment:N.Treatment (added Temp.Treatment, N.Treatment and CO2.Treatment by themselves too...)
    anova(end_flowers_preds) # Significant: Water.Treatment 
    anova(end_fruits_preds) # Significant: none, but each variable is included separately for kicks
    
    # Make ANOVA table out of predicts output
    start_leaves_preds_aov <- data.frame(Anova(start_leaves_preds)[3]); start_leaves_preds_aov <- start_leaves_preds_aov %>% mutate(Measure = rep("start"), Phase = rep("Leaves"))
    start_flowers_preds_aov <- data.frame(Anova(start_flowers_preds)[3]); start_flowers_preds_aov <- start_flowers_preds_aov %>% mutate(Measure = rep("start"), Phase = rep("OpenFlowers"))
    start_fruits_preds_aov <- data.frame(Anova(start_fruits_preds)[3]); start_fruits_preds_aov <- start_fruits_preds_aov %>% mutate(Measure = rep("start"), Phase = rep("RipeFruits"))
    
    end_leaves_preds_aov <- data.frame(Anova(end_leaves_preds)[3]); end_leaves_preds_aov <- end_leaves_preds_aov %>% mutate(Measure = rep("end"), Phase = rep("Leaves"))
    end_flowers_preds_aov <- data.frame(Anova(end_flowers_preds)[3]); end_flowers_preds_aov <- end_flowers_preds_aov %>% mutate(Measure = rep("end"), Phase = rep("OpenFlowers"))
    end_fruits_preds_aov <- data.frame(Anova(end_fruits_preds)[3]); end_fruits_preds_aov <- end_fruits_preds_aov %>% mutate(Measure = rep("end"), Phase = rep("RipeFruits"))
    
    # bind
    aov_preds <- bind_rows(start_leaves_preds_aov,
                           start_flowers_preds_aov,
                           start_fruits_preds_aov,
                           end_leaves_preds_aov,
                           end_flowers_preds_aov,
                           end_fruits_preds_aov)
    # add significance codes
    aov_preds <- aov_preds %>% mutate(Significance = ifelse(Pr..Chisq. < 0.001, "***",
                                                            ifelse(Pr..Chisq. < 0.01, "**", 
                                                                   ifelse(Pr..Chisq. < 0.05, "*", 
                                                                          ifelse(Pr..Chisq. < 0.1, ".", ""))))) %>%
      # create row names for join to preds dataframe (below)
      rownames_to_column(var = "rowname") %>%
      mutate(Treatment_Combo = case_when(
        grepl("Temp", rowname) ~ "Warming",
        grepl("Water", rowname) ~ "-W",
        grepl("N", rowname) ~ "+N",
        grepl("CO2", rowname) ~ "+C",
        TRUE ~ ""
      )) %>%
      mutate(Treatment_Combo = paste0(
        ifelse(grepl("Temp", rowname), "+T", ""),
        ifelse(grepl("Water", rowname), "-W", ""),
        ifelse(grepl("N", rowname), "+N", ""),
        ifelse(grepl("CO2", rowname), "+C", "")
      )) %>%
      column_to_rownames(var = "rowname") %>%
      mutate(Treatment_Combo = ifelse(Treatment_Combo == "+N+C", "+C+N", Treatment_Combo))
      
    # check out DFs via AICs
    AICc(start_leaves_preds,
         start_flowers_preds,
         start_fruits_preds,
         end_leaves_preds,
         end_flowers_preds,
         end_fruits_preds)
    # Unlike linear mixed models (lmer), generalized linear mixed models (glmer) do not directly compute denominator degrees of freedom (DF). This is because glmer uses maximum likelihood (ML) or restricted maximum likelihood (REML) estimation, which does not inherently assign DF to fixed effects.

    # However, there are several methods for estimating denominator DF that are commonly used:
    # Extract DF for fixed effects
    emm <- emmeans(start_leaves_preds, ~ Temp.Treatment)
    emm
    # "df = Inf" means that glmer is using Wald tests, which assume infinite DF.
    
# Step 3: Generate a New Datasets for Predictions
    # Create a new datasets for predictions
    newdat <- expand.grid(
          Temp.Treatment = c("HTamb", "HTelv"),
          Water.Treatment = c("H2Oamb", "H2Oneg"),
          CO2.Treatment = c("ambient CO2", "+CO2"),
          N.Treatment = c("ambient N", "+N")
        ) %>% # now we can tailor this dataset 
        mutate(
        # add treatment combo column
        Treatment_Combo = case_when(
          Temp.Treatment == "HTelv" & Water.Treatment == "H2Oamb" & CO2.Treatment == "ambient CO2" & N.Treatment == "ambient N" ~ "+T",
          Temp.Treatment == "HTamb" & Water.Treatment == "H2Oneg" & CO2.Treatment == "ambient CO2" & N.Treatment == "ambient N" ~ "-W",
          Temp.Treatment == "HTamb" & Water.Treatment == "H2Oamb" & CO2.Treatment == "+CO2" & N.Treatment == "ambient N" ~ "+C",
          Temp.Treatment == "HTamb" & Water.Treatment == "H2Oamb" & CO2.Treatment == "ambient CO2" & N.Treatment == "+N" ~ "+N",
          Temp.Treatment == "HTamb" & Water.Treatment == "H2Oamb" & CO2.Treatment == "ambient CO2" & N.Treatment == "ambient N" ~ "Ctrl", # All ambient
          TRUE ~ paste0(
            ifelse(Temp.Treatment == "HTelv", "+T", ""),
            ifelse(Water.Treatment == "H2Oneg", "-W", ""),
            ifelse(CO2.Treatment == "+CO2", "+C", ""),
            ifelse(N.Treatment == "+N", "+N", "")
          )
        )
      ) %>%
      # remove instances of 3 & 4-way interactions
      mutate(
        non_ambient_count = rowSums(data.frame(
          Temp.Treatment != "HTamb",
          Water.Treatment != "H2Oamb",
          CO2.Treatment != "ambient CO2",
          N.Treatment != "ambient N"
        ))
      ) %>%
      filter(non_ambient_count <= 2) %>% # Keep only rows with <= 2 non-ambient variables
      dplyr::select(-non_ambient_count) # Drop the helper column
    
# Step 4: Generate specific datasets for models
    # START OF LEAVES
    # Significant: Temp.Treatment, Temp.Treatment:N.Treatment (added N.Treatment by itself too...)
    newdat_start_leaves <- newdat %>%
      # filter to get specific instances we're testing with this response
      filter(Treatment_Combo %in% c("Ctrl", "+T", "+N", "+T+N"))
    
    # START OF FLOWERS
    # Significant: Temp.Treatment 
    newdat_start_flowers <- newdat %>%
      # filter to get specific instances we're testing with this response
      filter(Treatment_Combo %in% c("Ctrl", "+T"))
    
    # START OF FRUITS
    # Significant: Temp.Treatment, Water.Treatment
    newdat_start_fruits <- newdat %>%
      # filter to get specific instances we're testing with this response
      filter(Treatment_Combo %in% c("Ctrl", "+T", "-W"))
    
    # END OF LEAVES
    # Significant: Temp.Treatment:CO2.Treatment (added Temp.Treatment and CO2.Treatment by themselves too...)
    newdat_end_leaves <- newdat %>%
      # filter to get specific instances we're testing with this response
      filter(Treatment_Combo %in% c("Ctrl", "+T", "+C", "+N", "+C+N", "+T+C"))
    # END OF FLOWERS
    # Significant: Water.Treatment 
    newdat_end_flowers <- newdat %>%
      # filter to get specific instances we're testing with this response
      filter(Treatment_Combo %in% c("Ctrl", "+T", "-W", "+C", "+N"))
    
    # END OF FRUITS
    # Significant: none, but each variable is included separately for kicks
    newdat_end_fruits <- newdat %>%
      # filter to get specific instances we're testing with this response
      filter(Treatment_Combo %in% c("Ctrl", "+T", "-W", "+C", "+N"))  

# Step 5: Extract predictors and standard errors (RUN IN SEQUENCE!)
    # START OF LEAVES
    # Generate Model Matrix for New Data
    fixed_formula <- reformulate(attr(terms(start_leaves_preds), "term.labels"))  # Extract fixed effects formula
    X <- model.matrix(fixed_formula, newdat_start_leaves)  # Generate model matrix
    # Extract Fixed Effect Estimates and Variance-Covariance Matrix
    beta_hat <- fixef(start_leaves_preds)  # Fixed effects
    vcov_beta <- vcov(start_leaves_preds)  # Covariance matrix
    # Compute Predictions and Standard Errors
    link_predictions <- X %*% beta_hat  # Predictions on link scale
    link_se <- sqrt(diag(X %*% vcov_beta %*% t(X)))  # Standard errors
    # Store Predictions in Dataframe
    newdat_start_leaves1 <- newdat_start_leaves %>%
      mutate(
        link_se = exp(link_se),
        predicted = exp(link_predictions),  # Poisson uses log-link, so exp() transforms predictions back
        lower = exp(link_predictions - 1.96 * link_se),  # Confidence interval lower bound
        upper = exp(link_predictions + 1.96 * link_se),  # Confidence interval upper bound
        Measure = "start",  # Label for Measure
        Phase = "Leaves"  # Label for Phase
      )
      # Verify Predictions Using predict()
      newdat_start_leaves1 <- newdat_start_leaves1 %>%
        mutate(
          predict_check = predict(start_leaves_preds, newdata = newdat_start_leaves, type = "response", re.form = NA)
        )
        
    # START OF FLOWERS
    # Generate Model Matrix for New Data
    fixed_formula <- reformulate(attr(terms(start_flowers_preds), "term.labels"))  # Extract fixed effects formula
    X <- model.matrix(fixed_formula, newdat_start_flowers)  # Generate model matrix
    # Extract Fixed Effect Estimates and Variance-Covariance Matrix
    beta_hat <- fixef(start_flowers_preds)  # Fixed effects
    vcov_beta <- vcov(start_flowers_preds)  # Covariance matrix
    # Compute Predictions and Standard Errors
    link_predictions <- X %*% beta_hat  # Predictions on link scale
    link_se <- sqrt(diag(X %*% vcov_beta %*% t(X)))  # Standard errors
    # Store Predictions in Dataframe
    newdat_start_flowers1 <- newdat_start_flowers %>%
      mutate(
        link_se = exp(link_se),
        predicted = exp(link_predictions),  # Poisson uses log-link, so exp() transforms predictions back
        lower = exp(link_predictions - 1.96 * link_se),  # Confidence interval lower bound
        upper = exp(link_predictions + 1.96 * link_se),  # Confidence interval upper bound
        Measure = "start",  # Label for Measure
        Phase = "OpenFlowers"  # Label for Phase
      )
      # Verify Predictions Using predict()
      newdat_start_flowers1 <- newdat_start_flowers1 %>%
        mutate(
          predict_check = predict(start_flowers_preds, newdata = newdat_start_flowers, type = "response", re.form = NA)
        )
        
    # START OF FRUITS
    # Generate Model Matrix for New Data
    fixed_formula <- reformulate(attr(terms(start_fruits_preds), "term.labels"))  # Extract fixed effects formula
    X <- model.matrix(fixed_formula, newdat_start_fruits)  # Generate model matrix
    # Extract Fixed Effect Estimates and Variance-Covariance Matrix
    beta_hat <- fixef(start_fruits_preds)  # Fixed effects
    vcov_beta <- vcov(start_fruits_preds)  # Covariance matrix
    # Compute Predictions and Standard Errors
    link_predictions <- X %*% beta_hat  # Predictions on link scale
    link_se <- sqrt(diag(X %*% vcov_beta %*% t(X)))  # Standard errors
    # Store Predictions in Dataframe
    newdat_start_fruits1 <- newdat_start_fruits %>%
      mutate(
        link_se = exp(link_se),
        predicted = exp(link_predictions),  # Poisson uses log-link, so exp() transforms predictions back
        lower = exp(link_predictions - 1.96 * link_se),  # Confidence interval lower bound
        upper = exp(link_predictions + 1.96 * link_se),  # Confidence interval upper bound
        Measure = "start",  # Label for Measure
        Phase = "RipeFruits"  # Label for Phase
      )
    # Verify Predictions Using predict()
    newdat_start_fruits1 <- newdat_start_fruits1 %>%
      mutate(
        predict_check = predict(start_fruits_preds, newdata = newdat_start_fruits, type = "response", re.form = NA)
      )
      
    # END OF LEAVES
    # Extract the fixed-effects formula (no response variable, no random effects)
        fixed_formula <- reformulate(attr(terms(end_leaves_preds), "term.labels"))
        # Generate the model matrix for the new data
        X <- model.matrix(fixed_formula, newdat_end_leaves)
        # Extract fixed-effect estimates and covariance matrix
        beta_hat <- fixef(end_leaves_preds)  # Fixed effects
        vcov_beta <- vcov(end_leaves_preds)  # Covariance matrix of fixed effects
        # Calculate predictions on the link scale
        link_predictions <- X %*% beta_hat
        # Standard errors for the link scale predictions
        link_se <- sqrt(diag(X %*% vcov_beta %*% t(X)))
        # Generate predictions
        end_leaves_predictions <- predict(end_leaves_preds, newdata = newdat_end_leaves, type = "link", re.form = NA) 
        # Back-transform predictions and confidence intervals
        newdat_end_leaves1 <- newdat_end_leaves %>%
          mutate(
                link_se = (exp(link_se))^2,
                predicted = 365 - (exp(link_predictions))^2 - 1,  # Back-transform predictions
            Measure = "start",  # Label for Measure
            Phase = "Leaves"  # Label for Phase
          ) %>% 
          mutate(Measure = rep("end"), Phase = rep("Leaves")) #%>% dplyr::select(-c(se_sqrt,predicted_sqrt,lower_sqrt,upper_sqrt))
        # Verify Predictions Using predict()
        newdat_end_leaves1 <- newdat_end_leaves1 %>%
          mutate(
            predict_check = predict(end_leaves_preds, newdata = newdat_end_leaves, type = "response", re.form = NA)
          )
        
    # END OF FLOWERS
    # Extract the fixed-effects formula (no response variable, no random effects)
        fixed_formula <- reformulate(attr(terms(end_flowers_preds), "term.labels"))
        # Generate the model matrix for the new data
        X <- model.matrix(fixed_formula, newdat_end_flowers)
        # Extract fixed-effect estimates and covariance matrix
        beta_hat <- fixef(end_flowers_preds)  # Fixed effects
        vcov_beta <- vcov(end_flowers_preds)  # Covariance matrix of fixed effects
        # Calculate predictions on the link scale
        link_predictions <- X %*% beta_hat
        # Standard errors for the link scale predictions
        link_se <- sqrt(diag(X %*% vcov_beta %*% t(X)))
        # Generate predictions
        end_flowers_predictions <- predict(end_flowers_preds, newdata = newdat_end_flowers, type = "link", re.form = NA) 
        # Back-transform predictions and confidence intervals
        newdat_end_flowers1 <- newdat_end_flowers %>%
          mutate(
                link_se = (exp(link_se))^2,
                predicted = 365 - (exp(link_predictions))^2 - 1,  # Back-transform predictions
            Measure = "start",  # Label for Measure
            Phase = "Leaves"  # Label for Phase
          ) %>% 
          mutate(Measure = rep("end"), Phase = rep("OpenFlowers")) 
        # Verify Predictions Using predict()
        newdat_end_flowers1 <- newdat_end_flowers1 %>%
          mutate(
            predict_check = predict(end_flowers_preds, newdata = newdat_end_flowers, type = "response", re.form = NA)
          )
        
    # END OF FRUITS
    # Extract the fixed-effects formula (no response variable, no random effects)
        fixed_formula <- reformulate(attr(terms(end_fruits_preds), "term.labels"))
        # Generate the model matrix for the new data
        X <- model.matrix(fixed_formula, newdat_end_fruits)
        # Extract fixed-effect estimates and covariance matrix
        beta_hat <- fixef(end_fruits_preds)  # Fixed effects
        vcov_beta <- vcov(end_fruits_preds)  # Covariance matrix of fixed effects
        # Calculate predictions on the link scale
        link_predictions <- X %*% beta_hat
        # Standard errors for the link scale predictions
        link_se <- sqrt(diag(X %*% vcov_beta %*% t(X)))
        # Generate predictions
        end_fruits_predictions <- predict(end_fruits_preds, newdata = newdat_end_fruits, type = "link", re.form = NA) 
        # Back-transform predictions and confidence intervals
        newdat_end_fruits1 <- newdat_end_fruits %>%
          mutate(
                link_se = (exp(link_se))^2,
                predicted = 365 - (exp(link_predictions))^2 - 1,  # Back-transform predictions
            Measure = "start",  # Label for Measure
            Phase = "Leaves"  # Label for Phase
          ) %>% 
          mutate(Measure = rep("end"), Phase = rep("RipeFruits")) #%>% dplyr::select(-c(se_sqrt,predicted_sqrt,lower_sqrt,upper_sqrt))
        # Verify Predictions Using predict()
        newdat_end_fruits1 <- newdat_end_fruits1 %>%
          mutate(
            predict_check = predict(end_fruits_preds, newdata = newdat_end_fruits, type = "response", re.form = NA)
          )
        
        # Combine dataframes
        newdata_all_models <- bind_rows(newdat_start_leaves1, 
                                        newdat_start_flowers1, 
                                        newdat_start_fruits1, 
                                        newdat_end_leaves1, 
                                        newdat_end_flowers1, 
                                        newdat_end_fruits1)
        # Add anova significance codes
        newdata_all_models <- newdata_all_models %>%
          left_join(., aov_preds, by = c("Treatment_Combo", "Phase", "Measure"))
        
    
# Step 6: Plots
        newdata_all_models$Measure <- factor(newdata_all_models$Measure, levels = c("start", "end"))
        # labels for plots
        Phase.labs <- c("Leaves", "Open Flowers", "Ripe Fruits")
        names(Phase.labs) <- c("Leaves", "OpenFlowers", "RipeFruits")
         
        Measure.labs <- c("Start", "End")
        names(Measure.labs) <- c("start", "end")

newdata_all_models$Treatment_Combo <- factor(newdata_all_models$Treatment_Combo, levels = c("Ctrl", "+T", "+N", "-W", "+C", "+T+N", "+T+C","+C+N"))

# PLOT WITH ALL MODELS
ggplot(newdata_all_models, aes(y = Treatment_Combo, x = predicted, color = Treatment_Combo)) +
  geom_point(position = position_dodge(0.5), size = 3) +
  geom_errorbar(aes(xmin = predicted - link_se, xmax = predicted + link_se), position = position_dodge(0.5), width = 0.2) +
  labs(
    title = "Model Predictions for Phase Start and End",
    x = "Mean Day of Year ± 1 SE",
    y = "Treatment"
  ) +
  geom_shadowtext(aes(label = Significance),
                  position = position_dodge(width = 0.6),
                  hjust = -0.25, vjust = -0.25, size = 4,
                  color = "black", bg.color = "white", bg.r = 0.1, # Add white outline
                  show.legend = FALSE) + 
  theme(
    panel.border = element_rect(color = "black", fill = NA),
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = "grey90", size = 0.25),
    panel.grid.minor = element_blank(),
    strip.background.x=element_rect(color = NA,  fill=NA), 
    strip.background.y=element_rect(color = "black",  fill=NA),
    strip.text = element_text(size = 10, face = "bold"),
    plot.title = element_text(hjust = 0.5),
    axis.title = element_text(size = 12), #face = "bold"
    legend.position = "right",
    legend.background = element_rect(fill = "white"),
    legend.title = element_text()
  ) +
  scale_color_manual(values = c("black", 
                                "firebrick1", 
                                "palegreen3",
                                "cornflowerblue", 
                                "grey50", 
                                "gold3",
                                "chocolate4",
                                "darkgreen"), name = "Treatment") +
  facet_wrap(Phase ~ Measure, scales = "free_x", ncol = 2,
             labeller = labeller(Phase = Phase.labs, Measure = Measure.labs))

# INDIVIDUAL PLOTS FOR EACH MODEL
trt_palette <- c("Ctrl"="black", 
                "+T"="firebrick1", 
                "+N"="palegreen3",
                "-W"="cornflowerblue", 
                "+C"="grey50", 
                "+T+N"="gold3",
                "+T+C"="chocolate4",
                "+C+N"="darkgreen")
custom_theme <- theme(
  panel.border = element_rect(color = "black", fill = NA),
  panel.background = element_rect(fill = "white"),
  plot.background = element_rect(fill = "white"),
  panel.grid.major = element_line(color = "grey90", size = 0.25),
  panel.grid.minor = element_blank(),
  strip.background.x = element_rect(color = NA, fill = NA), 
  strip.background.y = element_rect(color = "black", fill = NA),
  strip.text = element_text(size = 10, face = "bold"),
  plot.title = element_text(face = "bold"),
  axis.title = element_text(size = 10), 
  legend.position = "right",
  legend.background = element_rect(fill = "white"),
  legend.title = element_text()
)

# Start of Leaves
newdata_all_models %>%
  filter(Measure == "start" & Phase == "Leaves") %>%
ggplot(., aes(y = Treatment_Combo, x = predicted, color = Treatment_Combo, linetype = Treatment_Combo)) +
  geom_point(position = position_dodge(0.5), size = 4) +
  geom_errorbar(aes(xmin = predicted - link_se, xmax = predicted + link_se), position = position_dodge(0.5), width = 0.1, size = 1) +
  labs(
    x = expression(bold("Start of Leaves") ~ "- Mean Day of Year ± 1 SE"),
    y = expression(bold("Treatment"))
  ) +
  custom_theme +
  scale_color_manual(values = trt_palette, name = " ", labels = c("Control", "Warming", "+Nitrogen", "Warming x +Nitrogen")) +
  scale_linetype_manual(
    values = c("Ctrl" = "solid", "+T" = "solid", "+N" = "dashed", "-W" = "dashed", "+T+N" = "solid"),
    name = " ", guide = "none"
    ) +
  guides(color = guide_legend(reverse = TRUE)) +
  scale_y_discrete(labels = c("Control", "Warming", "+Nitrogen", "Warming x +Nitrogen"))

# Start of flowers
newdata_all_models %>%
  filter(Measure == "start" & Phase == "OpenFlowers") %>%
ggplot(., aes(y = Treatment_Combo, x = predicted, color = Treatment_Combo, linetype = Treatment_Combo)) +
  geom_point(position = position_dodge(0.5), size = 4) +
  geom_errorbar(aes(xmin = predicted - link_se, xmax = predicted + link_se), position = position_dodge(0.5), width = 0.1, size = 1) +
  labs(
    x = expression(bold("Start of Open Flowers") ~ "- Mean Day of Year ± 1 SE"),
    y = expression(bold("Treatment"))
  ) +
  custom_theme +
  scale_color_manual(values = trt_palette, name = " ", labels = c("Control", "Warming")) +
  scale_linetype_manual(
    values = c("Ctrl" = "solid", "+T" = "solid"),
    name = " ",
    labels = c("Control", "Warming"),
  guide = "none") +
  guides(color = guide_legend(reverse = TRUE)) +
  scale_y_discrete(labels = c("Control", "Warming"))

# Start of fruits
newdata_all_models %>%
  filter(Measure == "start" & Phase == "RipeFruits") %>%
ggplot(., aes(y = Treatment_Combo, x = predicted, color = Treatment_Combo, linetype = Treatment_Combo)) +
  geom_point(position = position_dodge(0.5), size = 4) +
  geom_errorbar(aes(xmin = predicted - link_se, xmax = predicted + link_se), position = position_dodge(0.5), width = 0.1, size = 1) +
  labs(
    x = expression(bold("Start of Ripe Fruits") ~ "- Mean Day of Year ± 1 SE"),
    y = expression(bold("Treatment"))
  ) +
  custom_theme +
  scale_color_manual(values = trt_palette, name = " ", labels = c("Control", "Warming", "-Rainfall")) +
  scale_linetype_manual(
    values = c("Ctrl" = "solid", "+T" = "solid", "-W" = "solid"),
    name = " ",
  guide = "none") +
  guides(color = guide_legend(reverse = TRUE)) +
  scale_y_discrete(labels = c("Control", "Warming", "-Rainfall"))

# End of Leaves
newdata_all_models %>%
  filter(Measure == "end" & Phase == "Leaves") %>%
ggplot(., aes(y = Treatment_Combo, x = predicted, color = Treatment_Combo, linetype = Treatment_Combo)) +
  geom_point(position = position_dodge(0.5), size = 4) +
  geom_errorbar(aes(xmin = predicted - link_se, xmax = predicted + link_se), position = position_dodge(0.5), width = 0.1, size = 1) +
  labs(
    x = expression(bold("End of Leaf Senescence") ~ "- Mean Day of Year ± 1 SE"),
    y = expression(bold("Treatment"))
  ) +
  custom_theme +
  scale_color_manual(values = trt_palette, name = " ", labels = c("Control", "Warming", "+Nitrogen", bquote("+CO"[2]), bquote("Warming x +CO"[2]), bquote("+Nitrogen x +CO"[2]))) +
  scale_linetype_manual(
    values = c("Ctrl" = "solid", "+T" = "solid", "+N" = "dashed", "+C" = "dashed", "+T+C" = "solid", "+C+N" = "solid"),
    name = " ",
  guide = "none") +
  guides(color = guide_legend(reverse = TRUE)) +
  scale_y_discrete(labels = c("Control", "Warming", "+Nitrogen", bquote("+CO"[2]), bquote("Warming x +CO"[2]), bquote("+Nitrogen x +CO"[2])))

# End of flowers
ggplot(newdata_all_models %>%
         filter(Measure == "end" & Phase == "OpenFlowers"),
       aes(y = Treatment_Combo, x = predicted, color = Treatment_Combo, linetype = Treatment_Combo)) +
  geom_point(position = position_dodge(0.5), size = 4) +
  geom_errorbar(aes(xmin = predicted - link_se, xmax = predicted + link_se),
                position = position_dodge(0.5), width = 0.1, size = 1) +
  labs(
    x = expression(bold("End of Open Flowers") ~ "- Mean Day of Year ± 1 SE"),
    y = expression(bold("Treatment"))
  ) +
  custom_theme +
  scale_color_manual(
    values = trt_palette,
    name = " ",
    labels = c("Control", "Warming", "+Nitrogen", "-Rainfall", bquote("+CO"[2]))
  ) +
  scale_linetype_manual(
    values = c("Ctrl" = "solid", "+T" = "dashed", "+N" = "dashed", "-W" = "dashed", "+C" = "dashed"),
    name = " ",
    labels = c("Control", "Warming", "Nitrogen", "-Rainfall", bquote("+CO"[2])),
    guide = "none"
  ) +
  guides(color = guide_legend(reverse = TRUE)) +
  scale_y_discrete(labels = c("Control", "Warming", "+Nitrogen", "-Rainfall", bquote("+CO"[2])))

# End of fruits
newdata_all_models %>%
  filter(Measure == "end" & Phase == "RipeFruits") %>%
ggplot(., aes(y = Treatment_Combo, x = predicted, color = Treatment_Combo, linetype = Treatment_Combo)) +
  geom_point(position = position_dodge(0.5), size = 4) +
  geom_errorbar(aes(xmin = predicted - link_se, xmax = predicted + link_se), position = position_dodge(0.5), width = 0.1, size = 1) +
  labs(
    x = expression(bold("End of Ripe Fruits") ~ "- Mean Day of Year ± 1 SE"),
    y = expression(bold("Treatment"))
  ) +
  custom_theme +
  scale_color_manual(values = trt_palette, name = " ", labels = c("Control", "Warming", "+Nitrogen", "-Rainfall", bquote("+CO"[2]))) +
  scale_linetype_manual(
    values = c("Ctrl" = "solid", "+T" = "dashed", "+N" = "dashed", "-W" = "dashed", "+C" = "dashed"),
    name = " ",
    labels = c("Control", "Warming", "+Nitrogen", "-Rainfall", bquote("+CO"[2])),
   guide = "none") +
  guides(color = guide_legend(reverse = TRUE)) +
  scale_y_discrete(labels = c("Control", "Warming", "+Nitrogen", "-Rainfall", bquote("+CO"[2])))

```
[Back to Index](#index)

#### graph model coefficients, but include ALL non-sig. interactions {#all_trt_graphs}
```{r}
# Step 1: List the Models & Variables that are Significant in Each Model
    anova(start_leaves_glmer_poisson_4) 
    anova(start_flowers_glmer_poisson_4)
    anova(start_fruits_glmer_poisson_4) 
  
    anova(end_leaves_glmer_gamma_4)
    anova(end_flowers_glmer_gamma_4) 
    anova(end_fruits_glmer_gamma_4) 
    
    # check out DFs via AICs
    AICc(start_leaves_glmer_poisson_4,
         start_flowers_glmer_poisson_4,
         start_fruits_glmer_poisson_4,
         end_leaves_glmer_gamma_4,
         end_flowers_glmer_gamma_4,
         end_fruits_glmer_gamma_4)

# Step 2: Generate a New Datasets for Predictions
    # Create a new datasets for predictions
    newdat <- expand.grid(
          Temp.Treatment = c("HTamb", "HTelv"),
          Water.Treatment = c("H2Oamb", "H2Oneg"),
          CO2.Treatment = c("ambient CO2", "+CO2"),
          N.Treatment = c("ambient N", "+N")
        ) %>%  
        mutate(
        # add treatment combo column
        Treatment_Combo = case_when(
          Temp.Treatment == "HTelv" & Water.Treatment == "H2Oamb" & CO2.Treatment == "ambient CO2" & N.Treatment == "ambient N" ~ "+T",
          Temp.Treatment == "HTamb" & Water.Treatment == "H2Oneg" & CO2.Treatment == "ambient CO2" & N.Treatment == "ambient N" ~ "-W",
          Temp.Treatment == "HTamb" & Water.Treatment == "H2Oamb" & CO2.Treatment == "+CO2" & N.Treatment == "ambient N" ~ "+C",
          Temp.Treatment == "HTamb" & Water.Treatment == "H2Oamb" & CO2.Treatment == "ambient CO2" & N.Treatment == "+N" ~ "+N",
          Temp.Treatment == "HTamb" & Water.Treatment == "H2Oamb" & CO2.Treatment == "ambient CO2" & N.Treatment == "ambient N" ~ "Ctrl", # All ambient
          TRUE ~ paste0(
            ifelse(Temp.Treatment == "HTelv", "+T", ""),
            ifelse(Water.Treatment == "H2Oneg", "-W", ""),
            ifelse(CO2.Treatment == "+CO2", "+C", ""),
            ifelse(N.Treatment == "+N", "+N", "")
          )
        )
      ) 
    
# Step 3: Extract predictors and standard errors (RUN IN SEQUENCE!)
    # START OF LEAVES
    # Generate Model Matrix for New Data
    fixed_formula <- reformulate(attr(terms(start_leaves_glmer_poisson_4), "term.labels"))  # Extract fixed effects formula
    X <- model.matrix(fixed_formula, newdat)  # Generate model matrix
    # Extract Fixed Effect Estimates and Variance-Covariance Matrix
    beta_hat <- fixef(start_leaves_glmer_poisson_4)  # Fixed effects
    vcov_beta <- vcov(start_leaves_glmer_poisson_4)  # Covariance matrix
    # Compute Predictions and Standard Errors
    link_predictions <- X %*% beta_hat  # Predictions on link scale
    link_se <- sqrt(diag(X %*% vcov_beta %*% t(X)))  # Standard errors
    # Store Predictions in Dataframe
    newdat_start_leaves1 <- newdat %>%
      mutate(
        link_se = exp(link_se),
        predicted = exp(link_predictions),  # Poisson uses log-link, so exp() transforms predictions back
        lower = exp(link_predictions - 1.96 * link_se),  # Confidence interval lower bound
        upper = exp(link_predictions + 1.96 * link_se),  # Confidence interval upper bound
        Measure = "start",  # Label for Measure
        Phase = "Leaves"  # Label for Phase
      )
      # Verify Predictions Using predict()
      newdat_start_leaves1 <- newdat_start_leaves1 %>%
        mutate(
          predict_check = predict(start_leaves_glmer_poisson_4, newdata = newdat, type = "response", re.form = NA)
        )
        
    # START OF FLOWERS
    # Generate Model Matrix for New Data
    fixed_formula <- reformulate(attr(terms(start_flowers_glmer_poisson_4), "term.labels"))  # Extract fixed effects formula
    X <- model.matrix(fixed_formula, newdat)  # Generate model matrix
    # Extract Fixed Effect Estimates and Variance-Covariance Matrix
    beta_hat <- fixef(start_flowers_glmer_poisson_4)  # Fixed effects
    vcov_beta <- vcov(start_flowers_glmer_poisson_4)  # Covariance matrix
    # Compute Predictions and Standard Errors
    link_predictions <- X %*% beta_hat  # Predictions on link scale
    link_se <- sqrt(diag(X %*% vcov_beta %*% t(X)))  # Standard errors
    # Store Predictions in Dataframe
    newdat_start_flowers1 <- newdat %>%
      mutate(
        link_se = exp(link_se),
        predicted = exp(link_predictions),  # Poisson uses log-link, so exp() transforms predictions back
        lower = exp(link_predictions - 1.96 * link_se),  # Confidence interval lower bound
        upper = exp(link_predictions + 1.96 * link_se),  # Confidence interval upper bound
        Measure = "start",  # Label for Measure
        Phase = "OpenFlowers"  # Label for Phase
      )
      # Verify Predictions Using predict()
      newdat_start_flowers1 <- newdat_start_flowers1 %>%
        mutate(
          predict_check = predict(start_flowers_glmer_poisson_4, newdata = newdat, type = "response", re.form = NA)
        )
        
    # START OF FRUITS
    # Generate Model Matrix for New Data
    fixed_formula <- reformulate(attr(terms(start_fruits_glmer_poisson_4), "term.labels"))  # Extract fixed effects formula
    X <- model.matrix(fixed_formula, newdat)  # Generate model matrix
    # Extract Fixed Effect Estimates and Variance-Covariance Matrix
    beta_hat <- fixef(start_fruits_glmer_poisson_4)  # Fixed effects
    vcov_beta <- vcov(start_fruits_glmer_poisson_4)  # Covariance matrix
    # Compute Predictions and Standard Errors
    link_predictions <- X %*% beta_hat  # Predictions on link scale
    link_se <- sqrt(diag(X %*% vcov_beta %*% t(X)))  # Standard errors
    # Store Predictions in Dataframe
    newdat_start_fruits1 <- newdat %>%
      mutate(
        link_se = exp(link_se),
        predicted = exp(link_predictions),  # Poisson uses log-link, so exp() transforms predictions back
        lower = exp(link_predictions - 1.96 * link_se),  # Confidence interval lower bound
        upper = exp(link_predictions + 1.96 * link_se),  # Confidence interval upper bound
        Measure = "start",  # Label for Measure
        Phase = "RipeFruits"  # Label for Phase
      )
    # Verify Predictions Using predict()
    newdat_start_fruits1 <- newdat_start_fruits1 %>%
      mutate(
        predict_check = predict(start_fruits_glmer_poisson_4, newdata = newdat, type = "response", re.form = NA)
      )
      
    # END OF LEAVES
    # Extract the fixed-effects formula (no response variable, no random effects)
        fixed_formula <- reformulate(attr(terms(end_leaves_glmer_gamma_4), "term.labels"))
        # Generate the model matrix for the new data
        X <- model.matrix(fixed_formula, newdat)
        # Extract fixed-effect estimates and covariance matrix
        beta_hat <- fixef(end_leaves_glmer_gamma_4)  # Fixed effects
        vcov_beta <- vcov(end_leaves_glmer_gamma_4)  # Covariance matrix of fixed effects
        # Calculate predictions on the link scale
        link_predictions <- X %*% beta_hat
        # Standard errors for the link scale predictions
        link_se <- sqrt(diag(X %*% vcov_beta %*% t(X)))
        # Generate predictions
        end_leaves_predictions <- predict(end_leaves_glmer_gamma_4, newdata = newdat, type = "link", re.form = NA) 
        # Back-transform predictions and confidence intervals
        newdat_end_leaves1 <- newdat %>%
          mutate(
            link_se = (exp(link_se))^2,
            predicted = 365 - (exp(link_predictions))^2 - 1
          ) %>% mutate(Measure = rep("end_rev"), Phase = rep("Leaves"))
        # Verify Predictions Using predict()
        newdat_end_leaves1 <- newdat_end_leaves1 %>%
          mutate(
            predict_check = predict(end_leaves_glmer_gamma_4, newdata = newdat, type = "response", re.form = NA)
          )
        
    # END OF FLOWERS
    # Extract the fixed-effects formula (no response variable, no random effects)
        fixed_formula <- reformulate(attr(terms(end_flowers_glmer_gamma_4), "term.labels"))
        # Generate the model matrix for the new data
        X <- model.matrix(fixed_formula, newdat)
        # Extract fixed-effect estimates and covariance matrix
        beta_hat <- fixef(end_flowers_glmer_gamma_4)  # Fixed effects
        vcov_beta <- vcov(end_flowers_glmer_gamma_4)  # Covariance matrix of fixed effects
        # Calculate predictions on the link scale
        link_predictions <- X %*% beta_hat
        # Standard errors for the link scale predictions
        link_se <- sqrt(diag(X %*% vcov_beta %*% t(X)))
        # Generate predictions
        end_flowers_predictions <- predict(end_flowers_glmer_gamma_4, newdata = newdat, type = "link", re.form = NA) 
        # Back-transform predictions and confidence intervals
        newdat_end_flowers1 <- newdat %>%
          mutate(
            link_se = (exp(link_se))^2,
            predicted = 365 - (exp(link_predictions))^2 - 1
            # link_se = (exp(link_se))^2,
            # predicted = (exp(link_predictions))^2,  # Back-transform predictions
            # lower = (exp(link_predictions - 1.96 * link_se))^2,  # Lower bound
            # upper = (exp(link_predictions + 1.96 * link_se))^2   # Upper bound
          ) %>% mutate(Measure = rep("end_rev"), Phase = rep("OpenFlowers"))
        # Verify Predictions Using predict()
        newdat_end_flowers1 <- newdat_end_flowers1 %>%
          mutate(
            predict_check = predict(end_flowers_glmer_gamma_4, newdata = newdat, type = "response", re.form = NA)
          )
        
    # END OF FRUITS
    # Extract the fixed-effects formula (no response variable, no random effects)
        fixed_formula <- reformulate(attr(terms(end_fruits_glmer_gamma_4), "term.labels"))
        # Generate the model matrix for the new data
        X <- model.matrix(fixed_formula, newdat)
        # Extract fixed-effect estimates and covariance matrix
        beta_hat <- fixef(end_fruits_glmer_gamma_4)  # Fixed effects
        vcov_beta <- vcov(end_fruits_glmer_gamma_4)  # Covariance matrix of fixed effects
        # Calculate predictions on the link scale
        link_predictions <- X %*% beta_hat
        # Standard errors for the link scale predictions
        link_se <- sqrt(diag(X %*% vcov_beta %*% t(X)))
        # Generate predictions
        end_fruits_predictions <- predict(end_fruits_glmer_gamma_4, newdata = newdat, type = "link", re.form = NA) 
        # Back-transform predictions and confidence intervals
        newdat_end_fruits1 <- newdat %>%
          mutate(
            link_se = (exp(link_se))^2,
            predicted = 365 - (exp(link_predictions))^2 - 1
            # link_se = (exp(link_se))^2,
            # predicted = (exp(link_predictions))^2,  # Back-transform predictions
            # lower = (exp(link_predictions - 1.96 * link_se))^2,  # Lower bound
            # upper = (exp(link_predictions + 1.96 * link_se))^2   # Upper bound
          ) %>% mutate(Measure = rep("end_rev"), Phase = rep("RipeFruits"))
        # Verify Predictions Using predict()
        newdat_end_fruits1 <- newdat_end_fruits1 %>%
          mutate(
            predict_check = predict(end_fruits_glmer_gamma_4, newdata = newdat, type = "response", re.form = NA)
          )
        
        # Combine dataframes
        newdata_all_models <- bind_rows(newdat_start_leaves1, 
                                        newdat_start_flowers1, 
                                        newdat_start_fruits1, 
                                        newdat_end_leaves1, 
                                        newdat_end_flowers1, 
                                        newdat_end_fruits1)
        
# Step 4: Plots
        newdata_all_models$Measure <- factor(newdata_all_models$Measure, levels = c("start", "end_rev"))
        # labels for plots
        Phase.labs <- c("Leaves", "Open Flowers", "Ripe Fruits")
        names(Phase.labs) <- c("Leaves", "OpenFlowers", "RipeFruits")
         
        Measure.labs <- c("Start", "End")
        names(Measure.labs) <- c("start", "end_rev")
        
        # Define custom labels for the interactions
        N_CO2_labels <- c("ambient N.ambient CO2" = "Ctrl", 
                          "ambient N.+CO2" = "+C", 
                          "+N.ambient CO2" = "+N", 
                          "+N.+CO2" = "+C+N")
        
        Water_Temp_labels <- c("HTamb.H2Oamb" = "Ctrl", 
                               "HTelv.H2Oamb" = "+T", 
                               "HTamb.H2Oneg" = "-W", 
                               "HTelv.H2Oneg" = "+T-W")
        
        # Create new columns by combining the relevant columns
        newdata_all_models$N_CO2_Combined <- with(newdata_all_models, interaction(N.Treatment, CO2.Treatment, sep = "."))
        newdata_all_models$Water_Temp_Combined <- with(newdata_all_models, interaction(Temp.Treatment, Water.Treatment, sep = "."))
        
        # Replace the values in the new columns with the custom labels
        newdata_all_models$N_CO2_Combined <- N_CO2_labels[as.character(newdata_all_models$N_CO2_Combined)]
        newdata_all_models$Water_Temp_Combined <- Water_Temp_labels[as.character(newdata_all_models$Water_Temp_Combined)]
        
        # Reorder the levels of the new variables
        newdata_all_models$N_CO2_Combined <- factor(newdata_all_models$N_CO2_Combined, levels = c("Ctrl", "+C", "+N", "+C+N"))
        newdata_all_models$Water_Temp_Combined <- factor(newdata_all_models$Water_Temp_Combined, levels = c("Ctrl", "+T", "-W", "+T-W"))
        
        # Find the value of 'predicted' where 'N_CO2_Combined' == 'Ctrl' & 'Water_Temp_Combined' == 'Ctrl' for each group of Phase and Measure
        ctrl_predicted_values <- newdata_all_models[newdata_all_models$N_CO2_Combined == "Ctrl" & newdata_all_models$Water_Temp_Combined == "Ctrl", ]
        ctrl_predicted_values <- aggregate(predicted ~ Phase + Measure, data = ctrl_predicted_values, FUN = mean)
        
        # Merge the ctrl_predicted_values back to the main dataframe
        newdata_all_models <- merge(newdata_all_models, ctrl_predicted_values, by = c("Phase", "Measure"), suffixes = c("", "_ctrl"))
        
        # Plot with new columns and dashed vertical lines
        ggplot(newdata_all_models, aes(y = N_CO2_Combined, x = predicted, color = Water_Temp_Combined)) +
          geom_point(position = position_dodge(0.5), size = 3) +
          geom_errorbar(aes(xmin = predicted - link_se, xmax = predicted + link_se), position = position_dodge(0.5), width = 0.2) +
          geom_vline(aes(xintercept = V1), linetype = "dashed", color = "grey50") +
          labs(
            title = "Model Predictions for Phase Start and End",
            subtitle = "Dashed line indicates true control",
            x = "Mean DOY ± SE",
            y = "Treatment"
          ) +
          theme(
              panel.border = element_rect(color = "black", fill = NA),
              panel.background = element_rect(fill = "white"),
              plot.background = element_rect(fill = "white"),
              panel.grid.major = element_line(color = "grey90", size = 0.25),
              panel.grid.minor = element_blank(),
              strip.background.x=element_rect(color = NA,  fill=NA), 
              strip.background.y=element_rect(color = "black",  fill=NA),
              strip.text = element_text(size = 10, face = "bold"),
              plot.title = element_text(hjust = 0.5),
              plot.subtitle = element_text(hjust = 0.5),
              axis.title = element_text(size = 12), #face = "bold"
              axis.text.y = element_text(face = "bold", size = 10),
              legend.position = "right",
              legend.background = element_rect(fill = "white"),
              legend.title = element_text()
            ) +
          scale_x_continuous(labels = scales::number_format(accuracy = 1)) + #scheme
          scale_color_manual(values = c("black", "firebrick1", "cornflowerblue", "maroon4"), name = "Treatment", labels = c("Ambient", "+T", "-W", "+T-W")) +
          facet_wrap(Phase ~ Measure, scales = "free_x", ncol = 2,
           labeller = labeller(Phase = Phase.labs, Measure = Measure.labs)) 
```
[Back to Index](#index)

#### calculate heat treatments effect on growing degree days {#gdd}
    
    # look at comparison of years in study to average of 1963 - 2011 period (previous 48 years)
```{r}
# just show the raw data over time. We don't necessarily make assumptions that the treatments worked

head(weather)

# Assuming your dataset is already loaded as 'weather'
# Convert the DATE column to Date format
weather <- weather %>%
  mutate(Date = as.Date(Date, format = "%m/%d/%Y"))

# Exclude the year 2020 from the dataset (no data collected during COVID yr 1)
weather_filtered <- weather %>%
  filter(year(Date) != 2020) %>%
  filter(year(Date) > 2011 & year(Date) < 2024) %>%
  # change temp to degrees C and precip to mm
  dplyr::mutate(MaxTemp.degC. = (MaxTemp.degF.- 32)*5/9,
                MinTemp.degC. = (MinTemp.degF.- 32)*5/9,
                Precip.mm. = Precip.inches.*25.4)
  

# Define a function to calculate GDD
calculate_gdd <- function(MinTemp.degC., MaxTemp.degC., base_temp = 10) {
  MinTemp.degC. <- ifelse(MinTemp.degC. < base_temp, base_temp, MinTemp.degC.)
  MaxTemp.degC. <- ifelse(MaxTemp.degC. < base_temp, base_temp, MaxTemp.degC.)
  gdd <- ((MaxTemp.degC. + MinTemp.degC.) / 2) - base_temp
  return(max(gdd, 0))
}

# 0. Compare gdd and precip. accumulation rates 
weather_averages <- weather %>%
  # change temp to degrees C and precip to mm
  dplyr::mutate(MaxTemp.degC. = (MaxTemp.degF.- 32)*5/9,
                MinTemp.degC. = (MinTemp.degF.- 32)*5/9,
                Precip.mm. = Precip.inches.*25.4) %>%
  # filter to growing season
  dplyr::mutate(Month = format(Date, "%m")) %>%
  filter(Month %in% c("05", "06", "07", "08", "09")) %>%
  # filter to 1963 - 2011 data
  filter(year(Date) < 2011) %>% 
  # change temp to degrees C and precip to mm
  dplyr::mutate(MaxTemp.degC. = (MaxTemp.degF.- 32)*5/9,
                MinTemp.degC. = (MinTemp.degF.- 32)*5/9,
                Precip.mm. = Precip.inches.*25.4) %>%
  # calculate GDD
  mutate(GDD = mapply(calculate_gdd, MinTemp.degC., MaxTemp.degC.)) %>%
  # summarize by year initially...
  group_by(Year = year(Date)) %>%
  summarize(Cumulative_GDD = sum(GDD, na.rm = TRUE),
            Total_Precipitation_mm = sum(Precip.mm., na.rm = TRUE)
            ) %>% 
  arrange(Year) %>%
  # then take averages across timeframe
  ungroup() %>%
  dplyr::summarize(Cumulative_GDD_avg1963_2011 = mean(Cumulative_GDD),
                   Total_Precipitation_mm_avg1963_2011 = mean(Total_Precipitation_mm),
                   
                    n = n(),
        
                    mean_gdd = mean(Cumulative_GDD, na.rm = TRUE),
                    se_gdd = sd(Cumulative_GDD, na.rm = TRUE) / sqrt(n()),
        
                    mean_precip = mean(Total_Precipitation_mm, na.rm = TRUE),
                    se_precip = sd(Total_Precipitation_mm, na.rm = TRUE) / sqrt(n())
                   )
  

# Calculate GDD and GDD with temperature treatment for each day
weather_gdd <- weather_filtered %>%
  mutate(GDD = mapply(calculate_gdd, MinTemp.degC., MaxTemp.degC.)) 

# year effects
cumulative_gdd_year <- weather_gdd %>%
  group_by(Year = year(Date)) %>%
  summarize(Cumulative_GDD = sum(GDD, na.rm = TRUE),
            Total_Precipitation_mm = sum(Precip.mm., na.rm = TRUE)
            ) %>%
  arrange(Year) %>%
  mutate(GDD_diff_from_histAvg = Cumulative_GDD - weather_averages$Cumulative_GDD_avg1963_2011,
         Precip_diff_from_histAvg = Total_Precipitation_mm - weather_averages$Total_Precipitation_mm_avg1963_2011) %>%
  # Ensure Year is a factor with descending levels
  mutate(Year = factor(Year, levels = sort(unique(Year), decreasing = TRUE)))

# growing season effects
cumulative_gdd_gs <- weather_gdd %>%
  mutate(Month = month(Date)) %>%
  filter(Month %in% c(5, 6, 7, 8, 9)) %>%
  dplyr::select(-c(Month)) %>%
  group_by(Year = year(Date)) %>%
  summarize(Cumulative_GDD = sum(GDD, na.rm = TRUE),
            Total_Precipitation_mm = sum(Precip.mm., na.rm = TRUE) 
            ) %>%
  arrange(Year) %>%
  mutate(GDD_diff_from_histAvg = Cumulative_GDD - weather_averages$Cumulative_GDD_avg1963_2011,
         Precip_diff_from_histAvg = Total_Precipitation_mm - weather_averages$Total_Precipitation_mm_avg1963_2011) %>%
  # Ensure Year is a factor with descending levels
  mutate(Year = factor(Year, levels = sort(unique(Year), decreasing = TRUE)))


# View the resulting datasets
print("Cumulative GDD and Precipitation by Year")
print(cumulative_gdd_year)

# Create a dataset with cumulative GDD and total precipitation by month and year
cumulative_gdd_month_year <- weather_gdd %>%
  group_by(Year = year(Date), Month = month(Date)) %>%
  summarize(Cumulative_GDD = sum(GDD),
            #Cumulative_GDD_Treatment = sum(GDD_Temp.Treatment, na.rm = TRUE),
            Total_Precipitation_mm = sum(Precip.mm.) #,
            #Total_Precipitation = sum(PRCP_Water.Treatment, na.rm = TRUE)
            ) %>%
  arrange(Year, Month)

# View the resulting datasets
print("Cumulative GDD and Precipitation by Year")
print(cumulative_gdd_year)

# precip by year
cumulative_gdd_month_year_plot <- cumulative_gdd_month_year %>%
  mutate(YearCat = as.factor(Year)) %>%
  ungroup() %>%
  group_by(Year) %>%
  mutate(Cumulative_Precipitation = cumsum(Total_Precipitation_mm),
         Cumulative_GDD = cumsum(Cumulative_GDD)) %>%
  arrange(Month)

cumulative_gdd_month_year_plot <- cumulative_gdd_month_year_plot %>%
  filter(Month < 11)
# Plot A: Growing Degree Day Accumulation
gdd_accumulation_yr <- ggplot(cumulative_gdd_month_year_plot, aes(x = Month)) +
  geom_point(aes(y = Cumulative_GDD, col = YearCat), size = 3, shape = 16) +
  geom_line(aes(y = Cumulative_GDD, group = YearCat, col = YearCat), size = 1) +
  labs(title = "A",
       x = "Month", y = "Growing Degree Day Accumulation") +
  scale_color_viridis(option = "F", discrete = TRUE, name = "Year") +
  scale_x_discrete(limits = c("Jan","Feb","Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct")) +  
  theme_minimal() +
  theme(legend.position = "none",  # Show legend to differentiate years
        axis.text.x = element_text(angle = 45, hjust = 1))

# Plot C: Total Precipitation Accumulation
precip_accumulation_yr <- ggplot(cumulative_gdd_month_year_plot, aes(x = Month)) +
  geom_point(aes(y = Cumulative_Precipitation, col = YearCat), size = 3, shape = 16) +
  geom_line(aes(y = Cumulative_Precipitation, group = YearCat, col = YearCat), size = 1) +
  labs(title = "C",
       x = "Month", y = "Precipitation Accumulation (mm)") +
  scale_color_viridis(option = "G", discrete = TRUE, name = "Year") +
 scale_x_discrete(limits = c("Jan","Feb","Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct")) +
  theme_minimal() +
  theme(legend.position = "none",  # Show legend to differentiate years
        axis.text.x = element_text(angle = 45, hjust = 1))

# Plot cumulative GDD by Year
gdd_year_plot <- ggplot(cumulative_gdd_year, aes(x = factor(Year), y = Cumulative_GDD, fill = factor(Year))) +
  geom_bar(stat = "identity", alpha = 0.7, 
           col = "black",
           width = 0.5) +
  labs(
      title = "B",
       x = "Year", y = "Total Growing Degree Days",
       fill = "Year") +
  theme_minimal() +
  scale_fill_viridis(option = "F", discrete = TRUE, name = "Year") +
  theme(legend.position = "right",
        legend.title = element_text(size = 10),
        axis.text.x = element_text(angle = 45, hjust = 1))

# Plot cumulative precip by Year
precip_year_plot <- ggplot(cumulative_gdd_year, aes(x = factor(Year), y = Total_Precipitation_mm, fill = factor(Year))) +
  geom_bar(stat = "identity", alpha = 0.7,
           col = "black", width = 0.5) +
  labs(
       title = "D",
       x = "Year", y = "Total Growing Season Precipitation (mm)",
       fill = "Year") +
  theme_minimal() +
  scale_fill_viridis(option = "G", discrete = TRUE, name = "Year") + 
  theme(legend.position = "right",
        legend.title = element_text(size = 10),
        axis.text.x = element_text(angle = 45, hjust = 1))

# Display the plots
ggarrange(gdd_accumulation_yr,gdd_year_plot,precip_accumulation_yr,precip_year_plot) 

# SPEI graph
spei2$SPEI_4cat <- factor(spei2$SPEI_4cat, levels = c("Extremely wet", "Wet", "Normal", "Dry", "Extremely dry")) # formerly spei
spei2 %>% # formerly spei
  mutate(Year = as.numeric(Year)) %>%
  ggplot(., aes(x = factor(Year), y = SPEI04, fill = SPEI_4cat)) +
  geom_bar(stat = "identity", col = "black", width = 0.5) +
  labs(
    title = "E",
    x = "Year", y = "SPEI index value (May - August)",
    fill = "SPEI category") +
  geom_hline(yintercept = 0) +
  theme_minimal() +
  scale_fill_manual(values = c("Extremely wet" = "#1f78b4", "Wet" = "#a6cee3", "Normal" = "#ffffff", "Dry" = "#eaa082", "Extremely dry" = "#c62a15")) +
  theme(legend.position = "right",
        legend.title = element_text(size = 10),
        axis.text.x = element_text(angle = 45, hjust = 1)) + scale_x_discrete(limits = as.character(2012:2023))


# Calculate mean and standard error for GDD and GDD_Temperature by month
cumulative_gdd_summary <- cumulative_gdd_month_year %>%
  group_by(Month) %>%
  summarize(
    mean_GDD = mean(Cumulative_GDD, na.rm = TRUE),
    se_GDD = sd(Cumulative_GDD, na.rm = TRUE) / sqrt(n())
  )

# Plot difference from historical (1963 - 2011) average

# Extract reference values
gdd_ref <- weather_averages$Cumulative_GDD_avg1963_2011
precip_ref <- weather_averages$Total_Precipitation_mm_avg1963_2011

# Create GDD difference plot
gdd_plot <- ggplot(cumulative_gdd_gs, aes(x = GDD_diff_from_histAvg, y = Year, fill = GDD_diff_from_histAvg)) +
  geom_col(color = "black") +
  xlim(-300, 300) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  scale_y_discrete(limits = rev(levels(cumulative_gdd_gs$Year))) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  labs(
    x = "GDD Difference from Historical Average",
    y = "Year"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

gdd_plot_allyrs <- cumulative_gdd_gs %>%
  ungroup() %>%
  dplyr::select(-c(Year)) %>%
  mutate(Year = rep("    All")) %>%
  group_by(Year) %>%
  dplyr::summarize(GDD_diff_from_histAvg = mean(GDD_diff_from_histAvg)) %>%
  ggplot(., aes(x = GDD_diff_from_histAvg, y = Year)) +
  geom_col(color = "black",fill = "#FFF0F0") +
  xlim(-300, 300) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  labs(
    x = "Average GDD Difference, 2012 - 2023",
    y = ""
  ) +
  theme_minimal() +
  theme(legend.position = "none")

# Create Precipitation difference plot
precip_plot <- ggplot(cumulative_gdd_gs, aes(x = Precip_diff_from_histAvg, y = Year, fill = Precip_diff_from_histAvg)) +
  geom_col(color = "black") +
  xlim(-300, 300) +
  scale_fill_gradient2(low = "darkgoldenrod3", mid = "white", high = "turquoise3", midpoint = 0) +
  scale_y_discrete(limits = rev(levels(cumulative_gdd_gs$Year))) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  labs(
   # title = "Growing Season Precipitation Difference from Historical Average",
    x = "Precipitation Difference (mm) from Historical Average",
    y = ""
  ) +
  theme_minimal() +
  theme(legend.position = "none")

precip_plot_allyrs <- cumulative_gdd_gs %>%
  ungroup() %>%
  dplyr::select(-c(Year)) %>%
  mutate(Year = rep("    All")) %>%
  group_by(Year) %>%
  dplyr::summarize(Precip_diff_from_histAvg = mean(Precip_diff_from_histAvg)) %>%
  ggplot(., aes(x = Precip_diff_from_histAvg, y = Year)) +
  geom_col(color = "black",fill = "#F7F0E1") +
  xlim(-300, 300) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  labs(
    x = "Average Precipitation Difference (mm), 2012 - 2023",
    y = ""
  ) +
  theme_minimal() +
  theme(legend.position = "none")

# arrange all plots
grid.arrange(gdd_plot, precip_plot, gdd_plot_allyrs, precip_plot_allyrs, heights=c(0.83, 0.17), nrow=2)

# end climate info 2
```
[Back to Index](#index)

#### GDD by species {#sps_gdd}
```{r}
head(weather_gdd)

# calculate day of the year from weather dataset
weather_gdd2 <- weather_gdd %>%
  group_by(Year = year(Date)) %>%
  dplyr::mutate(doy = yday(Date),
                YearCat = as.character(Year),
                cumulative_GDD = cumsum(GDD),
                cumulative_precip = cumsum(Precip.mm.))

# Filter to get the lowest value of "start" per year for each combination of Species.observed, treatment, and cumulative_GDD
df_filtered_start <- df2 %>%
  dplyr::select(start, Year, Species.observed, CO2.Treatment, N.Treatment, Water.Treatment, Temp.Treatment, Phase) %>%
  group_by(Year, Species.observed, CO2.Treatment, N.Treatment, Water.Treatment, Temp.Treatment, Phase) %>%
  filter(CO2.Treatment == "ambient CO2",
         N.Treatment == "ambient N",
         Temp.Treatment == "HTamb",
         Water.Treatment == "H2Oamb") %>%
  summarize_all(mean) %>%
  ungroup() %>%
  left_join(., weather_gdd2[, c("doy","YearCat","cumulative_GDD","cumulative_precip")], by = c("Year"="YearCat","start"="doy"))

df_filtered_end <- df2 %>%
  dplyr::select(end, Year, Species.observed, CO2.Treatment, N.Treatment, Water.Treatment, Temp.Treatment, Phase) %>%
  group_by(Year, Species.observed, CO2.Treatment, N.Treatment, Water.Treatment, Temp.Treatment, Phase) %>%
  filter(CO2.Treatment == "ambient CO2",
         N.Treatment == "ambient N",
         Temp.Treatment == "HTamb",
         Water.Treatment == "H2Oamb") %>%
  summarize_all(mean) %>%
  ungroup() %>%
  left_join(., weather_gdd2[, c("doy","YearCat","cumulative_GDD","cumulative_precip")], by = c("Year"="YearCat","end"="doy"))

# Create the ggplot
            phase.labs <- c("Leaves", "Open Flowers", "Ripe Fruits")
            names(phase.labs) <- c("Leaves", "OpenFlowers", "RipeFruits")
start_ctrl <- ggplot(df_filtered_start, aes(y = cumulative_GDD, x = start, fill = Species.observed)) +
  geom_point(alpha = 0.5, size = 2, pch = 21) +
  facet_wrap(~ Phase, ncol = 1, scales = "free_y", labeller = labeller(Phase = phase.labs)) +
  labs(
       y = "Cumulative GDD",
       x = "Start (DOY)",
       color = "Species") +
  scale_fill_viridis(option = "H", discrete = T,
                     labels = c(expression(italic("Agropyron repens")),
                                expression(italic("Andropogon gerardii")),
                                expression(italic("Bromus inermis")),
                                expression(italic("Lespedeza capitata")),
                                expression(italic("Lupinus perennis")),
                                expression(italic("Poa pratensis")),
                                expression(italic("Schizachyrium scoparium")),
                                expression(italic("Solidago rigida")))) +
   theme(
    panel.border = element_rect(color = "black", fill = NA),
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = "grey90", size = 0.25),
    panel.grid.minor = element_blank(),
    strip.background.x=element_rect(color = NA,  fill=NA), 
    strip.background.y=element_rect(color = "black",  fill=NA),
    strip.text = element_text(size = 10, face = "bold"),
    plot.title = element_text(hjust = 0.5),
    axis.title = element_text(size = 12), 
    legend.position = "right",
    legend.text.align = 0,
    legend.background = element_rect(fill = "white"),
    legend.title = element_text()
  );start_ctrl
end_ctrl <- ggplot(df_filtered_end, aes(y = cumulative_GDD, x = end, fill = Species.observed)) +
  geom_point(alpha = 0.5, size = 2, pch = 21) +
  facet_wrap(~ Phase, ncol = 1, scales = "free_y", labeller = labeller(Phase = phase.labs)) +
  labs(
       y = "",
       x = "End (DOY)",
       color = "Species") +
  scale_fill_viridis(option = "H", discrete = T,
                     labels = c(expression(italic("Agropyron repens")),
                                expression(italic("Andropogon gerardii")),
                                expression(italic("Bromus inermis")),
                                expression(italic("Lespedeza capitata")),
                                expression(italic("Lupinus perennis")),
                                expression(italic("Poa pratensis")),
                                expression(italic("Schizachyrium scoparium")),
                                expression(italic("Solidago rigida")))) +
  theme(
    panel.border = element_rect(color = "black", fill = NA),
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = "grey90", size = 0.25),
    panel.grid.minor = element_blank(),
    strip.background.x=element_rect(color = NA,  fill=NA), 
    strip.background.y=element_rect(color = "black",  fill=NA),
    strip.text = element_text(size = 10, face = "bold"),
    plot.title = element_text(hjust = 0.5),
    axis.title = element_text(size = 12), #face = "bold"
    legend.position = "right",
    legend.text.align = 0,
    legend.background = element_rect(fill = "white"),
    legend.title = element_text()
  )
ggarrange(start_ctrl, end_ctrl, ncol = 2, common.legend = TRUE, legend = "right")
```
[Back to Index](#index)

#### main treatment effects (only) on individual species (regardless of year) {#sps_trt}
```{r}
# Step 1: Generate a New Datasets for Predictions
    # Create a new datasets for predictions
    newdat_sp <- expand.grid(
          Temp.Treatment = c("HTamb", "HTelv"),
          Water.Treatment = c("H2Oamb", "H2Oneg"),
          CO2.Treatment = c("ambient CO2", "+CO2"),
          N.Treatment = c("ambient N", "+N")
        ) %>% # now we can tailor this dataset 
        mutate(
        # add treatment combo column
        Treatment_Combo = case_when(
          Temp.Treatment == "HTelv" & Water.Treatment == "H2Oamb" & CO2.Treatment == "ambient CO2" & N.Treatment == "ambient N" ~ "+T",
          Temp.Treatment == "HTamb" & Water.Treatment == "H2Oneg" & CO2.Treatment == "ambient CO2" & N.Treatment == "ambient N" ~ "-W",
          Temp.Treatment == "HTamb" & Water.Treatment == "H2Oamb" & CO2.Treatment == "+CO2" & N.Treatment == "ambient N" ~ "+C",
          Temp.Treatment == "HTamb" & Water.Treatment == "H2Oamb" & CO2.Treatment == "ambient CO2" & N.Treatment == "+N" ~ "+N",
          Temp.Treatment == "HTamb" & Water.Treatment == "H2Oamb" & CO2.Treatment == "ambient CO2" & N.Treatment == "ambient N" ~ "Ctrl", # All ambient
          TRUE ~ paste0(
            ifelse(Temp.Treatment == "HTelv", "+T", ""),
            ifelse(Water.Treatment == "H2Oneg", "-W", ""),
            ifelse(CO2.Treatment == "+CO2", "+C", ""),
            ifelse(N.Treatment == "+N", "+N", "")
          )
        )
      ) %>%
      # remove instances of 3 & 4-way interactions
      mutate(
        non_ambient_count = rowSums(data.frame(
          Temp.Treatment != "HTamb",
          Water.Treatment != "H2Oamb",
          CO2.Treatment != "ambient CO2",
          N.Treatment != "ambient N"
        ))
      ) %>%
      filter(non_ambient_count <= 1) %>% # Keep only rows with <= 1 non-ambient variables
      dplyr::select(-non_ambient_count) # Drop the helper column

# Step 2: Define the levels for each factor
phases <- c("Leaves", "OpenFlowers", "RipeFruits")
measures <- c("start", "sqrt_end_rev")  # Updated measures
species <- c("agrre", "andge", "broin", "lesca", "luppe", "poapr", "schsc", "solri")

# Create an empty dataframe to store all results
newdata_all_models <- data.frame()

# Step 3: Loop through all combinations of phase, measure, and species
for (phase in phases) {
  for (measure in measures) {
    for (sp in species) {
      
      # (1) Subset data
      df_subset <- subset(df2, Phase == phase & Species.observed == sp)

      # Skip if no data available for this combination
      if (nrow(df_subset) == 0) next

      # (2) Assign the correct family
      if (measure == "sqrt_end_rev") {
        family_type <- Gamma(link = "log")
      } else {
        family_type <- poisson()
      }

      # (3) Fit the model
      model <- glmer(
        as.formula(paste0(measure, " ~ Temp.Treatment + Water.Treatment + CO2.Treatment + N.Treatment + 
                           (1 | ExpYrCat) + (1 | Ring:Plot)")),
        data = df_subset,
        family = family_type
      )

      # (4) Generate Model Matrix for New Data
      fixed_formula <- reformulate(attr(terms(model), "term.labels"))  
      X <- model.matrix(fixed_formula, newdat_sp)

      # Extract Fixed Effect Estimates and Variance-Covariance Matrix
      beta_hat <- fixef(model)  
      vcov_beta <- vcov(model)  

      # Compute Predictions and Standard Errors
      link_predictions <- X %*% beta_hat  
      link_se <- sqrt(diag(X %*% vcov_beta %*% t(X)))  

      # (5) Adjust transformation based on measure type
      if (measure == "sqrt_end_rev") {
        newdat_temp <- newdat_sp %>%
          mutate(
            link_se = (exp(link_se))^2,
            predicted = 365 - (exp(link_predictions))^2 - 1,
            Measure = measure,
            Phase = phase,
            Species = sp
          )
      } else {  # Default Poisson transformations for "start"
        newdat_temp <- newdat_sp %>%
          mutate(
            link_se = exp(link_se),
            predicted = exp(link_predictions),
            lower = exp(link_predictions - 1.96 * link_se),
            upper = exp(link_predictions + 1.96 * link_se),
            Measure = measure,
            Phase = phase,
            Species = sp
          )
      }

      # (6) Verify Predictions Using predict()
      newdat_temp <- newdat_temp %>%
        mutate(
          predict_check = predict(model, newdata = newdat_sp, type = "response", re.form = NA)
        )

      # Append to the main dataframe
      newdata_all_models <- bind_rows(newdata_all_models, newdat_temp)
    }
  }
}

# View final combined dataframe
head(newdata_all_models) 

# Step 4: Plotz
        newdata_all_models$Measure <- factor(newdata_all_models$Measure, levels = c("start", "sqrt_end_rev"))
        # labels for plots
        Phase.labs <- c("Leaves", "Open Flowers", "Ripe Fruits")
        names(Phase.labs) <- c("Leaves", "OpenFlowers", "RipeFruits")
         
        Measure.labs <- c("Start", "End")
        names(Measure.labs) <- c("start", "sqrt_end_rev")

# Custom vectors for the new labels
newdata_all_models$Species <- factor(newdata_all_models$Species, levels = c("andge","broin","agrre","lesca","luppe","poapr","schsc","solri"))
 
species_labels <- c(
  andge = expression(italic("A. gerardii")),
  broin = expression(italic("B. inermis")),
  agrre = expression(italic("E. repens")),
  lesca = expression(italic("L. capitata")),
  luppe = expression(italic("L. perennis")),
  poapr = expression(italic("P. pratensis")),
  schsc = expression(italic("S. scoparium")),
  solri = expression(italic("S. rigida"))
)

treatment_colors <- c(
  "Ctrl" = "black",        # Black
  "+T" = "firebrick1", # Dark red
  "+C" = "grey50",  # Grey
  "-W" = "cornflowerblue",# Medium blue
  "+N" = "palegreen3"     # Medium sea green
)

# Reorder TrtCombo levels
newdata_all_models$Treatment_Combo <- factor(newdata_all_models$Treatment_Combo, levels = c("Ctrl", "+T", "-W", "+C", "+N"))

# Create model/response column
newdata_all_models$Model0 <- paste(newdata_all_models$Phase, newdata_all_models$Measure, sep = " ")

# Reorder model levels
newdata_all_models <- newdata_all_models %>%
  mutate(Model = ifelse(Model0 == "Leaves start", "Start of Leaves",
                        ifelse(Model0 == "OpenFlowers start", "Start of Flowers",
                               ifelse(Model0 == "RipeFruits start", "Start of Fruits",
                                      ifelse(Model0 == "Leaves sqrt_end_rev", "End of Leaves",
                                             ifelse(Model0 == "OpenFlowers sqrt_end_rev", "End of Flowers",
                                                    ifelse(Model0 == "RipeFruits sqrt_end_rev", "End of Fruits", Model0)))))))

newdata_all_models$Model <- factor(newdata_all_models$Model , levels = c("Start of Leaves", "Start of Flowers", "Start of Fruits", "End of Leaves", "End of Flowers", "End of Fruits"))

# Plotting
pacman::p_load(patchwork)

newdata_all_models %>%
  split(~Species) %>%  # Split by Species
  lapply(function(.data) {
    x_title <- unique(.data$Species)

    # Remove strip labels for all but the last facet
    remove_strip_x <- if (x_title != last(unique(newdata_all_models$Species))) 
      theme(strip.text.x = element_blank()) else theme()

    ggplot(.data, aes(y = Treatment_Combo, x = predicted, color = Treatment_Combo)) +
      geom_point(position = position_dodge(0.75), size = 1.5) +
      geom_errorbar(aes(xmin = predicted - link_se, xmax = predicted + link_se), 
                    position = position_dodge(0.5), width = 0) +
      #labs(y = x_title) +  # Move label to y-axis instead of strip
      ylab(species_labels[[x_title]]) +
      facet_wrap(
        facets = vars(Model),  # Facet by Model
        scales = "free_x",  # Allow different x-scales per facet
        nrow = 1,
        strip.position = "bottom",
        labeller = labeller(Species = as_labeller(species_labels))
      ) +
      annotate("segment", x = -Inf, xend = Inf, y = -Inf, yend = -Inf) +
      annotate("segment", x = -Inf, xend = -Inf, y = -Inf, yend = Inf) +
      scale_color_manual(values = treatment_colors) +
      theme_minimal() +
      theme(
        panel.grid.minor = element_blank(),
        panel.border = element_rect(colour = "black", fill = NA, size = 0.5),
        plot.subtitle = element_text(hjust = 0.5), 
        strip.text = element_text(size = 10, face = "bold"),
        legend.position = "none"
      ) +
      remove_strip_x  # Apply strip text removal conditionally
  }) %>%
  wrap_plots(ncol = 1) &  # Stack plots vertically  
  theme(
    strip.text = element_text(size = 10, face = "bold"),  # Facet label (strip) text size
    axis.text.x = element_text(size = 7, angle = 90),
    axis.text.y = element_text(size = 8),  # Y-axis tick text size
    axis.title.x = element_blank(),
    legend.title = element_blank(),
    panel.spacing = unit(2, "lines"),
    strip.placement = "outside",
    strip.switch.pad.wrap = unit(0, "lines")
  )
```
[Back to Index](#index)

#### year effects on individual species (regardless of treatment) {#sps_yr}
```{r}
df_raw_sp_yr <- df2 %>%
  ungroup() %>%
  dplyr::select(Year, Species.observed, Phase, start, end) %>%
  mutate(Phase = ifelse(Phase == "Leaves", "leaves",
                        ifelse(Phase == "OpenFlowers", "flowers",
                               ifelse(Phase == "RipeFruits", "fruits", NA)))) %>%
  rename(Start = start,
         End = end) %>%
  pivot_longer(col = c(Start:End), names_to = "Measure", values_to = "doy") %>%
  mutate(Response = paste(Measure, Phase, sep = " of ")) %>%
  dplyr::select(-c(Phase, Measure)) %>%
  group_by(Species.observed, Response, Year) %>%
  summarize_all(mean, na.rm = TRUE) %>%
  pivot_wider(names_from = Response, values_from = doy) %>% #%>% filter(Year != "2012")
  as.data.frame()
  
write.csv(df_raw_sp_yr, "YearEffects_PhaseMeasure_inc2023.csv")


# Plotting
pacman::p_load(patchwork)

df_raw_sp_yr %>%
  pivot_longer(cols = `End of flowers`:`Start of leaves`, names_to = "Response", values_to = "DOY") %>%
  split(~Species.observed) %>%  # Split by Species
  lapply(function(.data) {
    x_title <- unique(.data$Species.observed)

    # Remove strip labels for all but the last facet
    remove_strip_x <- if (x_title != last(unique(df_raw_sp_yr$Species.observed))) 
      theme(strip.text.x = element_blank()) else theme()

    ggplot(.data, aes(y = Year, x = DOY, color = "black")) +
      geom_point(position = position_dodge(0.75), size = 1.5) + 
      stat_summary(fun.data = mean_se,  geom = "errorbar", width = 0, size = 0.5, position = position_dodge(width = 0.5)) +
      stat_summary(fun=mean,geom="point", size=0.25,position = position_dodge(width = 0.5)) +
      # geom_errorbar(aes(xmin = predicted - link_se, xmax = predicted + link_se), 
      #               position = position_dodge(0.5), width = 0.2) +
      labs(y = x_title) +  # Move label to y-axis instead of strip
      facet_wrap(
        facets = vars(Response),  # Facet by Model
        scales = "free_x",  # Allow different x-scales per facet
        nrow = 1,
        strip.position = "bottom",
        labeller = labeller(Species = as_labeller(species_labels, label_parsed))
      ) +
      annotate("segment", x = -Inf, xend = Inf, y = -Inf, yend = -Inf) +
      annotate("segment", x = -Inf, xend = -Inf, y = -Inf, yend = Inf) +
      scale_color_manual(values = treatment_colors) +
      theme_minimal() +
      theme(
        panel.grid.minor = element_blank(),
        panel.border = element_rect(colour = "black", fill = NA, size = 0.5),
        plot.subtitle = element_text(hjust = 0.5), 
        strip.text = element_text(size = 10, face = "bold"),
        legend.position = "none"
      ) +
      remove_strip_x  # Apply strip text removal conditionally
  }) %>%
  wrap_plots(ncol = 1) &  # Stack plots vertically  
  theme(
    strip.text = element_text(size = 10, face = "bold"),  # Facet label (strip) text size
    axis.text.x = element_text(size = 7, angle = 90), # , hjust = 0.7
    axis.text.y = element_text(size = 8),  # Y-axis tick text size
    axis.title.x = element_blank(),
    legend.title = element_blank(),
    panel.spacing = unit(2, "lines"),
    strip.placement = "outside",
    strip.switch.pad.wrap = unit(0, "lines")
  )
```
[Back to Index](#index)

#### soil moisture plots {#soil_m}
*Note: this chunk takes a long time to run*
```{r}
head(sm)

# wrangle data
sm9 <- sm %>%
  # filter
  filter(CountOfSpecies == 9) %>%
  # reorganize date
  mutate(
        Date = as.Date(Date, "%m/%d/%Y"),
        Year = as.numeric(format(Date, "%Y")),
        Month = as.numeric(format(Date, "%m")),
        Day = as.numeric(format(Date, "%d"))) %>%
  dplyr::select(Date, Year, Month, Day, Ring, Plot, CountOfSpecies, Temp.Treatment, Water.Treatment, CO2.Treatment, N.Treatment, Method, Proportion.Soil.Moisture..0.17.cm., Proportion.Soil.Moisture..0.20.cm., Proportion.Soil.Moisture..22.39.cm.) %>%
  # pivot
  pivot_longer(cols = c(Proportion.Soil.Moisture..0.17.cm., Proportion.Soil.Moisture..0.20.cm., Proportion.Soil.Moisture..22.39.cm.), names_to = "depth", values_to = "prop_soil_m") %>%
  mutate(depth = ifelse(depth == "Proportion.Soil.Moisture..0.17.cm.", "0.17.cm",
                 ifelse(depth == "Proportion.Soil.Moisture..0.20.cm.", "0.20.cm",
                 ifelse(depth == "Proportion.Soil.Moisture..22.39.cm.", "22.39.cm", NA)))) %>%
  # filter
  filter(prop_soil_m != 0) %>%
  filter(!is.na(prop_soil_m)) %>%
  filter(Year >= 2012) %>%
  filter(Temp.Treatment != "") %>%
  # fix bugs
  mutate(N.Treatment = ifelse(N.Treatment == "NNenrichrich", "+N", 
                              ifelse(N.Treatment == "Namb", "ambient N", "")),
         CO2.Treatment = ifelse(CO2.Treatment == "Cenrich", "+CO2",
                                ifelse(CO2.Treatment == "Camb", "ambient CO2", ""))) %>%
  as.data.frame() %>%
  mutate(ExpYrCat = as.factor(-1*(2011-Year)))

a<-sm9 %>%
  ungroup() %>%
  group_by(ExpYrCat) %>%
  dplyr::select(ExpYrCat, Temp.Treatment, Water.Treatment, prop_soil_m, Year, Month) %>%
  pivot_wider(names_from = c("Temp.Treatment","Water.Treatment"), values_from = "prop_soil_m", values_fn = mean) %>%
  filter(!is.na(HTamb_)) %>%
  # probably a better way to do this...
  mutate(`-Water` = -(HTamb_H2Oamb - HTamb_H2Oneg),
         `+Heat` = -(HTamb_H2Oamb - HTelv_H2Oamb),
         `+Heat-Water` = -(HTamb_H2Oamb - HTelv_H2Oneg)) %>%
  dplyr::select(ExpYrCat, Year, Month, `-Water`, `+Heat`, `+Heat-Water`) %>%
  pivot_longer(cols = c(`-Water`, `+Heat`, `+Heat-Water`), names_to = "Treatment_effect", values_to = "soil_moisture_prop")

sm9 %>%
  ungroup() %>%
  group_by(ExpYrCat) %>%
  dplyr::select(ExpYrCat, Temp.Treatment, Water.Treatment, prop_soil_m, Year, Month) %>%
  pivot_wider(names_from = c("Temp.Treatment","Water.Treatment"), values_from = "prop_soil_m", values_fn = mean) %>%
  filter(!is.na(HTamb_)) %>%
  mutate(Month = ifelse(Month == 4, "Apr",
                        ifelse(Month == 5, "May", 
                               ifelse(Month == 6, "Jun", 
                                      ifelse(Month == 7, "Jul", 
                                             ifelse(Month == 8, "Aug",
                                                    ifelse(Month == 9, "Sep", NA)))))),
         Month = factor(Month, levels = c("Apr", "May", "Jun", "Jul", "Aug", "Sep"))) %>%
  # probably a better way to do this...
  mutate(`-Water` = -(HTamb_H2Oamb - HTamb_H2Oneg),
         `+Heat` = -(HTamb_H2Oamb - HTelv_H2Oamb),
         `+Heat-Water` = -(HTamb_H2Oamb - HTelv_H2Oneg)) %>%
  dplyr::select(ExpYrCat, Year, Month, `-Water`, `+Heat`, `+Heat-Water`) %>%
  pivot_longer(cols = c(`-Water`, `+Heat`, `+Heat-Water`), names_to = "Treatment_effect", values_to = "soil_moisture_prop") %>%
ggplot(., aes(x = Month, y = soil_moisture_prop, col = Treatment_effect)) +
  geom_point(alpha = 0.1) +
  scale_color_manual(values = c("cornflowerblue", "firebrick1", "maroon4"), name = "Treatment effect") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  stat_summary(fun.data = mean_se,  geom = "errorbar", width = 0, size = 0.75, position = position_dodge(width = 0.5)) +
  stat_summary(fun=mean,geom="point", size=2,position = position_dodge(width = 0.5)) +
  theme_minimal() +
  labs(x="Month", y = "Change in soil moisture prop. from ambient")

################################################################################
# run models
################################################################################
# april
sm9_apr <- lmer(prop_soil_m ~ Temp.Treatment * Water.Treatment + (1|Ring:Plot) + (1|depth) + (1|Year), data = subset(sm9, Month == 4)); anova(sm9_apr)
summary(sm9_apr)

# may
sm9_may <- lmer(prop_soil_m ~ Temp.Treatment * Water.Treatment + (1|Ring:Plot) + (1|depth) + (1|Year), data = subset(sm9, Month == 5)); anova(sm9_may)
summary(sm9_may)

# june
sm9_jun <- lmer(prop_soil_m ~ Temp.Treatment * Water.Treatment + (1|Ring:Plot) + (1|depth) + (1|Year), data = subset(sm9, Month == 6)); anova(sm9_jun)
summary(sm9_jun)

# july
sm9_jul <- lmer(prop_soil_m ~ Temp.Treatment * Water.Treatment + (1|Ring:Plot) + (1|depth) + (1|Year), data = subset(sm9, Month == 7)); anova(sm9_jul)
summary(sm9_jul)

# august
sm9_aug <- lmer(prop_soil_m ~ Temp.Treatment * Water.Treatment + (1|Ring:Plot) + (1|depth) + (1|Year), data = subset(sm9, Month == 8)); anova(sm9_aug)
summary(sm9_aug)

# september
sm9_sep <- lmer(prop_soil_m ~ Temp.Treatment * Water.Treatment + (1|Ring:Plot) + (1|depth) + (1|Year), data = subset(sm9, Month == 9)); anova(sm9_sep)
summary(sm9_sep)

################################################################################

# Required libraries
pacman::p_load(lme4,
               car,
               emmeans,
               dplyr)

# Initialize empty list to store results
all_results <- list()

# Loop over years 2012 to 2022
for (yr in 2012:2022) {
  
  # Fit model for current year
  model <- lmer(prop_soil_m ~ Temp.Treatment + (1|Ring/Plot) + (1|depth) + (1|Month),
                data = subset(sm9, Year == yr))
  
  # Extract Anova table
  aov_table <- Anova(model)
  aov_df <- as.data.frame(aov_table)
  aov_df$Effect <- rownames(aov_df)
  rownames(aov_df) <- NULL
  
  print("aov_df"); print(aov_df)
  
  # Filter only Temp.Treatment row and rename
  aov_extract <- aov_df %>%
    filter(Effect == "Temp.Treatment") %>%
    dplyr::select(Effect, Df, `Pr(>Chisq)`) %>%
    rename(p_value = `Pr(>Chisq)`)
  
  print("aov_extract"); print(aov_extract)
  
  # Extract emmeans and SE
  emm <- summary(emmeans(model, ~ Temp.Treatment))
  
  print("emm"); print(emm)
  
  # Combine emmeans output with Anova output
  result_df <- emm %>%
    mutate(Effect = "Temp.Treatment",
           Df = aov_extract$Df[1],
           p_value = aov_extract$p_value[1],
           Year = yr) %>%
    dplyr::select(Year, Effect, Temp.Treatment, emmean, SE, Df, p_value)
  
   print("result_df"); print(result_df)
  
  # Append to results list
  all_results[[as.character(yr)]] <- result_df
}

# Combine all results into one dataframe
final_df_1 <- bind_rows(all_results)

# View or export final result
print(final_df_1)

final_df %>%
  split(~Year) %>%  # Split by Species
  lapply(function(.data) {
    x_title <- unique(.data$Year)

    # Remove strip labels for all but the last facet
    remove_strip_x <- if (x_title != last(unique(final_df$Year))) 
      theme(strip.text.x = element_blank()) else theme()

    ggplot(.data, aes(y = Temp.Treatment, x = emmean, color = "black")) +
      geom_point(position = position_dodge(0.75), size = 1.5) +
      geom_errorbar(aes(xmin = emmean - SE, xmax = emmean + SE), 
                    position = position_dodge(0.5), width = 0.2) +
      labs(y = x_title) +  # Move label to y-axis instead of strip
      annotate("segment", x = -Inf, xend = Inf, y = -Inf, yend = -Inf) +
      annotate("segment", x = -Inf, xend = -Inf, y = -Inf, yend = Inf) +
      scale_color_manual(values = treatment_colors) +
      theme_minimal() +
      theme(
        panel.grid.minor = element_blank(),
        panel.border = element_rect(colour = "black", fill = NA, size = 0.5),
        plot.subtitle = element_text(hjust = 0.5), 
        strip.text = element_text(size = 10, face = "bold"),
        legend.position = "none"
      ) +
      remove_strip_x  # Apply strip text removal conditionally
  }) %>%
  wrap_plots(ncol = 1) &  # Stack plots vertically  
  theme(
    strip.text = element_text(size = 10, face = "bold"),  # Facet label (strip) text size
    axis.text.x = element_text(size = 7, angle = 90), # , hjust = 0.7
    axis.text.y = element_text(size = 8),  # Y-axis tick text size
    axis.title.x = element_blank(),
    legend.title = element_blank(),
    panel.spacing = unit(2, "lines"),
    strip.placement = "outside",
    strip.switch.pad.wrap = unit(0, "lines")
  )

# end soil moisture calculations
```

# here's a garden if you stayed until the end
```{r}
    # Define the Unicode emojis
    daisy <- "\U0001F33C"
    tulip <- "\U0001F337"
    cactus <- "\U0001F335"
    camellia <- "\U0001F33A"
    fern <- "\U0001F33F"
    sunflower <- "\U0001F33B"
    data <- "\U0001F4BE"
    
    # Print the emojis
    cat(daisy, tulip, cactus, camellia, fern, sunflower, data, "\n")

```
